---
title: "Example of Methylome Analysis with MethylIT using Cancer Datasets"
subtitle: <h1>A [Methyl-IT](https://github.com/genomaths/MethylIT) user guide. 
           A guide for users</h>
author: |
 | Robersy Sanchez$^1$
 | rus547@psu.edu
 |
 | Sally Mackenzie$^{1,2}$
 | sam795@psu.edu
 |
 | $^1$Department of Biology. 
 | Pennsylvania State University, University Park, PA 16802
 |
 | $^2$Department of Plant Science. 
 | Pennsylvania State University, University Park, PA 16802
date: "`r format(Sys.time(), '%d %B %Y')`"
fontsize: 11pt
fontfamily: "serif"
bibliography: bibliography.bib
output:
  rmarkdown::html_document: 
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: false
    number_sections: true
    highlight: tango
    theme: united
    geometry: margin=1in
    df_print: paged
    mathjax: local
    self_contained: false
    keep_md: yes
vignette: >
  %\VignetteIndexEntry{Cancer Example}
  %\VignetteEngine{knitr::knitr}
  \usepackage[utf8]{inputenc}
  
abstract: |
  An example of methylation analysis with Methyl-IT, a novel methylome analysis
  procedure based on a signal-detection and machine-learning approach, is
  provided here. Methylation analysis involves a signal detection problem, and
  the method was designed to discriminate methylation regulatory signal from
  background noise induced by thermal fluctuations. Methyl-IT enhances the
  resolution of genome methylation behavior to reveal network-associated
  responses, offering resolution of gene pathway influences not attainable with
  previous methods. Herein, an example of MethylIT application to the analysis
  of breast cancer methylomes is presented.
---


<style type="text/css">

.list-group-item.active, .list-group-item.active:focus, 
.list-group-item.active:hover {
    z-index: 2;
    color: #fff;
    background-color: #337ab7;
    border-color: #337ab7;
}

.tocify-subheader .tocify-item {
  font-size: 0.80em;
  padding-left: 25px;
  text-indent: 0;
}

.tocify-header .tocify-item {
  font-size: 0.85em;
  padding-left: 20px;
  text-indent: 0;
}

body{ /* Normal  */
      font-size: 18px;
      font-family: "Times New Roman", Times, serif;
      text-align: justify
  }
td {  /* Table  */
  font-size: 8px;
}

h1.title {
  font-size: 38px;
  font-family: "Times New Roman", Times, serif;
  color: DarkRed;
  .subTitle {
  font-size: 24px;
  font-family: "Times New Roman", Times, serif;
  color: DarkRed;
  }
}

h1 { /* Header 1 */
  font-size: 28px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
}
h2 { /* Header 2 */
    font-size: 22px;
    color: DarkBlue;
    font-family: "Times New Roman", Times, serif;
}
h3 { /* Header 3 */
   font-size: 18px;
   color: DarkBlue;
   font-family: "Times New Roman", Times, serif;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(options(width = 80))
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# [Methyl-IT](https://github.com/genomaths/MethylIT)
MethylIT is an R package for methylome analysis based on a signal-detection
machine learning approach (SD-ML). This approach is postulated to provide
greater sensitivity for resolving true signal from the methylation background
within the methylome [@Sanchez2016; @Sanchez2019]. Because the biological signal
created within the dynamic methylome environment characteristic of plants is not
free from background noise, the approach, designated Methyl-IT, includes the
application of signal detection theory [@Greiner2000; @Carter2016; @Harpaz2013;
@Kruspe2017]. A basic requirement for the application of signal detection is the
knowledge of the probability distribution of the background noise, which is used
as null hypothesis in the detection of the methylation signal. This probability
distribution is, in general, a member of generalized gamma distribution family,
and it can be deduced on a statistical mechanical/thermodynamics basis from DNA
methylation induced by thermal fluctuations [@Sanchez2016].

Herein, we provide an example of Methyl-IT application to the analysis of breast
cancer methylomes. Due to the size of human methylome the current example only
covers the analysis of chromosome 13. A full description of Methyl-IT
application of methylome analysis in plants is given in the manuscript
[@Sanchez2019].

The R packages required in this example are:
```{r libraries, message=FALSE}
library(MethylIT)
library(ggplot2) # graphic
library(reshape2) # To reshape the data frame
library(grid) # For multiple plots
library(gridExtra) # For multiple plots
library(VennDiagram) 
library(rtracklayer) # To import gene annotation
```

## Available datasets and reading
Methylome datasets of whole-genome bisulfite sequencing (WGBS) are available at
Gene Expression Omnibus (GEO DataSets). For the current example, datasets from
breast tissues (normal and cancer) and embryonic stem cells will be downloaded
from GEO. The data set are downloaded providing the GEO accession numbers for
each data set to the function 'getGEOSuppFiles' (see the function
[getGEOSuppFiles](https://genomaths.github.io/methylit/reference/getGEOSuppFiles.html)
help). Some times the ftp site is not available and the information must be
manually downloaded from GEO database.

```{r GEO, eval = TRUE, message=FALSE}

# Embryonic stem cells datasets
esc.files = getGEOSuppFiles(GEO = c("GSM2041690", "GSM2041691", "GSM2041692"),
                            verbose = FALSE)
# Breast tissues (normal, cancer, metastasis)
cancer.files = getGEOSuppFiles(GEO = c("GSM1279517", "GSM1279514",
                                       "GSM1279513"), verbose = FALSE)

```

The file path and name of each downloaded dataset is found in the output
variables 'esc.files' and 'cancer.files'.

## _Reading datasets_
Datasets for our example can be read with function [readCounts2GRangesList](https://genomaths.github.io/methylit/reference/readCounts2GRangesList.html).
To specify the reading of only chromosome 13, we can specify the parameter
'chromosomes = "Chr13"'. The symbol chromosome 13, in this case "Chr13", must
be consistent with the annotation provided in the given GEO dataset. Each file
is wholly read with the setting 'chromosomes = "Chr13"' and then the GRanges
are built only with chromosome 13, which could be time consuming. However,
users working on Linux OS can specify the reading of specific lines from each
file by using regular expressions. 

  For example, if only chromosomes 1 and 3 are required, then we can set
chromosomes = NULL (default) and 'chromosome.pattern = "^Chr[1,3]"'. This will
read all the lines in the downloaded files starting with the words "Chr1" or
"Chr3". If we are interested in chromosomes 1 and 2, then we can set
'chromosome.pattern = "^Chr[1-2]"'. If all the chromosomes are required, then
set chromosomes = NULL and chromosome.pattern = NULL (default). Obviously,
before read the files, user must check them to see which annotation was used by
the experimenters to denote the chromosomes, e.g., "chr13" or just "13", etc.

```{r reading datset, include = TRUE}
# Embryonic stem cells datasets
ref = readCounts2GRangesList(filenames = esc.files,  
                             sample.id = c("ESC1","ESC2", "ESC3"),
                             columns = c(seqnames = 1, start = 2,
                                         mC = 4, uC = 5), pattern = "^chr13",
                             remove = TRUE, verbose = FALSE)
# Breast tissues (normal, cancer, metastasis)
LR = readCounts2GRangesList(filenames = cancer.files,
                            sample.id = c("Breast_normal", "Breast_cancer",
                                          "Breast_metastasis"),
                            columns = c(seqnames = 1, start = 2,
                                         mC = 3, uC = 4),
                            remove = TRUE, pattern = "^13",
                            chromosome.names = "chr13", verbose = FALSE)
```

In real data analysis, a bigger sample size must be used. For the purpose of this
example, to illustrate the application of methylation analysis with Methyl-IT,
we will just add an artificial normal sample.

```{r virtual samples}
LR$Breast_normal1 <- poolFromGRlist(c(ref[1], LR$Breast_normal,
                                      LR$Breast_normal, LR$Breast_normal,
                                      LR$Breast_normal, LR$Breast_normal),
                                    stat = "mean", num.cores = 6L, verbose = FALSE)
```

In the metacolumn of the last GRanges object, mC and uC stand for the
methylated and unmethylated read counts, respectively. Notice that option 
'remove = TRUE' remove the decompressed files (default: FALSE, see 
?readCounts2GRangesList for more details about this function).

## The reference individual
Any two objects can be compared based on some measured variable if, and only if,
the measurements where taken in the same metric space, the same coordinate
system, in respect to the same origin of coordinates. Usually, in our daily 3D
experience, our brain automatically sets up the origin of coordinates equal to
zero. The differences found in the comparison depend on the reference used to
perform the measurements and from the metric system. The space where the objects
are located (or the set of objects) together with the metric is called metric
space.

Unfortunately, this is not the case for the methylation process. Each individual
from the same population and even each chromosome follows an independent
stochastic methylation process. This is not a surprise, since cytosine
methylation affect the mechanical properties of DNA molecule with strong
dependence of the DNA sequence context (until, at least. 6 bases at both sides
of each cytosine site) [@Severin2011; @Ngo2016].

To evaluate the methylation differences between individuals from control and
treatment we introduce a metric in the bidimensional space of methylation levels
$P_i = (p_i, 1-p_i)$. Vectors $P_i$ provide a measurement of the uncertainty of
methylation levels [@Sanchez2019]. However, to perform the comparison between
the uncertainty of methylation levels from each group of individuals, control
$(c)$ and treatment $(t)$, we should estimate the uncertainty variation with
respect to the same individual reference on the mentioned metric space. The
reason to measure the uncertainty variation with respect to the same reference
resides in that even sibling individuals follow an independent ontogenetic
development and, consequently, their developments follow independent stochastic
processes. This a consequence of the "omnipresent" action of the second law of
thermodynamics in living organisms. In the current example, we will create the
reference individual by pooling the methylation counts from the embryonic stem
cells.

It should be noticed that the results are sensitive to the reference used. The
statistics mean, median, or sum of the read counts at each cytosine site of
some control samples can be used to create a virtual reference sample. A virtual 
reference individual can be built with function [poolFromGRlist](https://genomaths.github.io/methylit/reference/poolFromGRlist.html).
It is up to the user whether to apply the 'row sum', 'row mean' or 'row median' of
methylated and unmethylated read counts at each cytosine site across
individuals. Notice that when "*mean*" is selected the virtual reference individual
is the group centroid, which play a fundamental role in multivariate statistic.

```{r pooling, include = TRUE}
Ref = poolFromGRlist(ref, stat = "mean", num.cores = 6L, verbose = FALSE)

Ref
```

Only direct lab experiments can reveal whether differences detected with
distinct references outside the experimental conditions for control and
treatment groups are real. The best reference would be estimated using a subset
of individuals from control group. Such a reference will contribute to remove
the intragroup variation, in control and in treatment groups, induced by
environmental changes external to or not controlled by the experimental
conditions.

Methylation analysis for each cytosine position is frequently performed in the
bidimensional space of $(methylated, unmethylated)$ read counts. Frequently,
Fisher's exact test is applied to a single cytosine position, under the null
hypothesis that the proportions
$p_{ct}=methylated_{ct}/(methylated_{ct}+unmethylated_{ct})$ and
$p_{tt}=methylated_{tt}/(methylated_{tt}+unmethylated_{tt})$ are the same for
control and treatment, respectively. In this case, the implicit reference point
for the counts at every cytosine positions is 
$(methylated = 0,unmethylated = 0)$, which corresponds to the point 
$P_i = (0, 1)$.

In our case, the Hellinger divergence (one of the metric used, here) of each
individual in respect to the reference is the variable to test in place of
$(methylated, unmethylated)$ read counts or the methylation levels
$P_i = (p_i, 1-p_i)$.

The use of references is restricted by the thermodynamics basis of the theory.
The current information-thermodynamics based approach is supported on the
following postulate:

  *"High changes of Hellinger divergences are less frequent than low changes,
  provided that the divergence is proportional to the amount of energy required
  to process the change of information in the methylation system"*.

The last postulate acknowledges the action of the second law of thermodynamics
on the biomolecular methylation system. For the methylation system, it implies
that the frequencies of the information divergences between methylation levels
must be proportional to a Boltzmann factor (see supplementary information from
reference [@Sanchez2016]). In other words, the frequencies of information
divergences values should follow a trend proportional to an exponential decay.
If we do not observe such a behavior, then either the reference is too far
from experimental condition or we are dealing with an extreme situation where
the methylation machinery in the cell is dysfunctional. The last situation is
found, for example, in the silencing mutation at the gene of
cytosine-DNA-methyltransferase in Arabidopsis *thaliana*. Methylation of
5-methylcytosine at CpG dinucleotides is maintained by MET1 in plants.

## Hellinger divergence estimation
To perform the comparison between the uncertainty of methylation levels
from each group of individuals, control $(c)$ and treatment $(t)$, the
divergence between the  methylation levels of each individual is estimated with
respect to the same reference on the metric space formed by the vector set 
$P_i = (p_i, 1-p_i)$ and the Hellinger divergence $H$. Basically, the 
information divergence between the methylation levels of an individual $j$ and
reference sample $r$ is estimated according to the Hellinger divergence given 
by the  formula: 
$$H(\hat p_{ij},\hat p_{ir}) = w_i[(\sqrt{\hat p_{ij}} - \sqrt{\hat
p_{ir}})^2+(\sqrt{1-\hat p_{ij}} - \sqrt{1-\hat p_{ir}})^2]$$ where 
$w_i = 2 \frac{m_{ij} m_{ir}}{m_{ij} + m_{ir}}$, 
$m_{ij} = {n_i}^{mC_j}+{n_i}^{uC_j}+1$, $m_{ir} = {n_i}^{mC_r}+{n_i}^{uC_r}+1$ 
and $j \in {\{c,t}\}$. This equation for Hellinger divergence is given in
reference [@Basu2010], but other information theoretical divergences can be
used as well. Next, the information divergence for control (Breast_normal) and
treatment (Breast_cancer and Breast_metastasis) samples are estimated with
respect to the reference virtual individual. A Bayesian correction of counts
can be selected or not. In a Bayesian framework, methylated read counts are
modeled by a beta-binomial distribution, which accounts for both the
biological and sampling variations [@Hebestreit2013; @Robinson2014;
@Dolzhenko2014]. In our case we adopted the Bayesian approach suggested in
reference [@Baldi2001] (Chapter 3). In a Bayesian framework with uniform
priors, the methylation level can be defined as: $p = (mC + 1)/(mC + uC + 2)$. 

However, the most natural statistical model for replicated BS-seq DNA
methylation measurements is beta-binomial (the beta distribution is a prior
conjugate of binomial distribution). We consider the parameter $p$ (methylation
level) in the binomial distribution as randomly drawn from a beta distribution.
The hyper-parameters $\alpha$ and $\beta$ from the beta-binomial distribution
are interpreted as pseudo-counts. The information divergence is estimated here
using the function [estimateDivergence](https://genomaths.github.io/methylit/reference/estimateDivergence.html):

```{r HD, include = TRUE}
HD = estimateDivergence(ref = Ref, indiv = LR, Bayesian = TRUE,  
                        min.coverage = 5, high.coverage = 300, 
                        percentile = 0.999, num.cores = 6L, tasks = 0L,
                        verbose = FALSE)

HD$Breast_cancer

```

Function 'estimateDivergence' returns a list of GRanges objects with the four 
columns of counts, the information divergence, and additional columns: 

  1. The original matrix of methylated ($c_i$) and unmethylated ($t_i$) read
     counts from control ($i=1$) and treatment ($i=2$) samples.
  2. "p1" and "p2": methylation levels for control and treatment, respectively.
  3. "bay.TV": total variation TV = p2 - p1.
  4. "TV": total variation based on simple counts: $TV=c1/(c1+t1)-c2/(c2+t2)$.
  5. "hdiv": Hellinger divergence.

If Bayesian = TRUE, results are based on the posterior estimations of
methylation levels $p1$ and $p2$. Filtering by coverage is provided at this
step which would be used unless previous filtering by coverage had been
applied. This is a pairwise filtering. Cytosine sites with 'coverage' >
'min.coverage' and 'coverage' < 'percentile' (e.g., 99.9 coverage percentile)
in at least one of the samples are preserved. The coverage percentile used is
the maximum estimated from both samples: reference and individual.

For some GEO datasets only the methylation levels for each cytosine site are
provided. In this case, Hellinger divergence can be estimated as given in
reference [@Sanchez2016]:
$$H(\hat p_{ij},\hat p_{ir}) = 2 \Big[(\sqrt{\hat p_{ij}} - \sqrt{\hat
p_{ir}})^2+(\sqrt{1-\hat p_{ij}} - \sqrt{1-\hat p_{ir}})^2\Big]$$

## _Descriptive statistical analysis_

A descriptive analysis on the distribution of Hellinger divergences of
methylation levels is recommended. Like in any other statistical analysis,
descriptive statistical analysis help us to detect potential issues in the raw
data. It is the user responsibility to perform quality check of his/her dataset
before start to apply Methyl-IT. Nevertheless, the quality checking of the raw
data is not perfect. So, it is healthy to sees for potential issues.

### _Histogram and boxplots of divergences estimated in each sample_
First, the data of interest (Hellinger divergences, "hdiv") are selected from
the GRanges objects:
```{r data4graphic, include = TRUE}
normal = HD$Breast_normal[, "hdiv"]
normal1 = HD$Breast_normal1[, "hdiv"]
cancer = HD$Breast_cancer[, "hdiv"]
metastasis = HD$Breast_metastasis[, "hdiv"]

```

Next, a single GRanges object is built from the above set of GRanges objects
using the function [uniqueGRanges](https://genomaths.github.io/methylit/reference/uniqueGRanges.html). 
Notice that the number of cores to use for parallel computation can be specified.
```{r hd4graphic, include = TRUE}
hd = uniqueGRanges(list(normal, normal1, cancer, metastasis), missing = NA,
                   verbose = FALSE, num.cores = 6L)
colnames(mcols(hd)) <- c("normal", "normal1", "cancer", "metastasis")
hd
```

Now, the Hellinger divergences estimated for each sample are in a single matrix
on the metacolumn of the GRanges object and we can proceed to build the
histogram and boxplot graphics for these data.

Finally, the data included in the graphic are:
```{r fig_data}
# Define an auxiliar function
fun_length <- function(x){
  return(data.frame(y = median(x) + 1, label = paste0("n = ", length(x))))
}

data <- data.frame(normal = hd$normal, normal1 = hd$normal1, 
                   cancer = hd$cancer, metastasis = hd$metastasis)
data = suppressMessages(melt(data))
colnames(data) <- c("Breast.tissue", "HD")
data = data[data$HD > 0, ]
DataFrame(data)
```


```{r fig1, fig.height = 7, fig.width = 20, message=FALSE}

p1 = ggplot(data, aes(x = HD, fill = Breast.tissue, colour = Breast.tissue)) + 
  geom_histogram(alpha = 0.5, binwidth = 0.5, position = "identity", na.rm = TRUE,
                 size = 0.7) +  xlim(-0.4,20) +
  theme(axis.title.x = element_text(face = "bold", size = 20),
        axis.text.x = element_text(face = "bold", size = 20, color = "black",
                                   hjust = 0.5, vjust = 0.75),
        axis.text.y = element_text(face = "bold", size = 20, color = "black"),
        axis.title.y = element_text(face = "bold", size = 20,color = "black"),
        legend.text = element_text(size = 20, face = "bold"),
        legend.title = element_text(size = 20, face = "bold")
        ) +
  ylab( "Counts" ) + 
  ggtitle("Histogram of Hellinger Divergence")

# For visualization purposes HD is limited to the interval 0 to 50
dt = data[ which(data$HD < 50), ] 
p2 = ggplot(dt,aes(x = Breast.tissue, y = HD , fill = Breast.tissue)) + 
  geom_boxplot(na.rm = TRUE)  +  ylim(-0.1, 20) +
  stat_summary(fun.data = fun_length, geom = "text",
               position = position_dodge(width = 0.9), vjust = 1,
               size = 6, fontface = "bold") +
  theme(axis.title.x = element_text(face = "bold", size = 20),
        axis.text.x = element_text(face = "bold", size = 20, color = "black",
                                   hjust = 0.5, vjust = 0.75),
        axis.text.y = element_text(face = "bold", size = 20, color = "black"),
        axis.title.y = element_text(face = "bold", size = 20,color = "black"),
        legend.position = "none"
        ) +
  ggtitle("Boxplot of Hellinger Divergence")
grid.arrange(p1, p2, ncol = 2)

```

Except for the tail, most of the methylation changes occurred under the area
covered by the density curve corresponding to the normal breast tissue. This is
theoretically expected. This area is explainable in statistical physical terms
and, theoretically, it should fit a Weibull distribution. The tails regions
cover the methylation changes that, with high probability, are not induced by
thermal fluctuation and are not addressed to stabilize the DNA molecule. These
changes are methylation signal. Professor David J. Miller (Department of
Electrical Engineering, Penn State) proposed modeling the distribution as a
mixed Weibull distribution to simultaneously describe the background
methylation noise and the methylation signal (personal communication, January,
2018). This model approach seems to be supported by the above histogram, but it
must be studied before being incorporated in a future version of Methyl-IT.

### _Checking cytosine read counts_

According to the parameter values set for the function 
[estimateDivergence](https://genomaths.github.io/methylit/reference/estimateDivergence.html),
for each cytosine site, at least one individual must have a 10 reads or more, and the 
maximum coverage must be 500 or less.
```{r coverage}

# Maximum coverage at each cytosine site
covr <- lapply(HD, function(x) {
     cov1 <- x$c1 + x$t1
     cov2 <- x$c2 + x$t2
     cov <- apply(cbind(cov1, cov2), 1, max)
     return(cov)
})

# Quantiles
do.call(rbind, lapply(covr, function(x) {
  q60 <- quantile(x, 0.6)
  q9999 <- quantile(x, 0.9999)
  idx1 <- which(x >= q60)
  idx2 <- which(x <= 500)
  q95 <- quantile(x, 0.95)

  idx <- intersect(idx1, idx2)
  return(c(round(summary(x)), q60, quantile(x, c(0.95, 0.99, 0.999, 0.9999)), 
           num.siteGreater_8 = sum(x >= 8), q60_to_500 = sum((x >= q60) & (x <= 500)),
           num.siteGreater_500 = sum(x > 500)))})
)

```

The quantiles indicates acceptable values for the coverages. Descriptive
statistics for methylated reads can be checked as well.

```{r meth-counts2}

# Maximum coverage at each cytosine site
methc <- lapply(HD, function(x) {
     r <- apply(cbind(x$c1, x$c2), 1, max)
     return(r)
})

# Quantiles
do.call(rbind, lapply(methc, function(x) {
    q10 <- quantile(x, 0.1)
    q60 <- quantile(x, 0.6)
    q9999 <- quantile(x, 0.9999)
    idx1 <- which(x >= q60)
    idx2 <- which(x <= 500)
    q95 <- quantile(x, 0.95)

    idx <- intersect(idx1, idx2)
    return(c(round(summary(x)), q10, q60, 
             quantile(x, c(0.95, 0.99, 0.999, 0.9999)), 
             num.siteGreater_8 = sum(x >= 8), 
             q60_to_500 = sum((x >= q60) & (x <= 500)),
             num.siteGreater_500 = sum(x > 500)))})
)

```

The critical values from empirical cumulative probability distributions will be
used in the downstream analysis.

```{r descriptive}

critical.val <- do.call(rbind, lapply(HD, function(x) {
  hd.95 = quantile(x$hdiv, 0.95)
  tv.95 = quantile(abs(x$bay.TV), 0.95)
  return(c(tv = tv.95, hd = hd.95))
}))

critical.val

```

## Nonlinear fit of probability distribution models
A basic requirement for the application of signal detection is the knowledge of
the probability distribution of the background noise. Probability distribution,
as a Weibull distribution model, can be deduced on a statistical
mechanical/thermodynamics basis for DNA methylation induced by thermal
fluctuations [@Sanchez2016]. Assuming that this background methylation
variation is consistent with a Poisson process, it can be distinguished from
variation associated with methylation regulatory machinery, which is
non-independent for all genomic regions [@Sanchez2016]. An
information-theoretic divergence to express the variation in methylation
induced by background thermal fluctuations will follow a Weibull distribution
model, provided that it is proportional to the minimum energy dissipated per
bit of information associated with the methylation change. The nonlinear fit to
a Weibull distribution model is performed by the function
[nonlinearFitDist](https://genomaths.github.io/methylit/reference/nonlinearFitDist.html).

In general, an information divergence of methylation levels will follow a
probability distribution from the family of generalized gamma distribution.
Next, the best fitted model will be selected from one of the following
distributions: two- and three-parameter Weibull distributions, gamma with two-
and three-parameters ("Gamma2P" or "Gamma3P"), and generalized gamma with
three-parameter ("GGamma3P").

The best model is selected on the basis of Akaike's information criterion and
the correlation coefficient of cross-validations for the nonlinear regressions
(R.Cross.val). These criteria evaluate different information inferred from the
models. AIC deals with the trade-off between the goodness of fit (GOF) of the
model and the complexity of the model, while R.Cross.val provides information on
the prediction power/performance of the model when confronted with external
dataset. Cross-validations for the nonlinear regressions (R.Cross.val) were
performed as described in reference [@Stevens2009]. In addition, Stein's formula
for adjusted R squared ($\rho$) was used as an estimator of the average
cross-validation predictive power [@Stevens2009].

In general, the best fitted model yields the best DMP classification performance
(see below). However, on special dataset, as the used in the current example.
Function
[gofReport](https://genomaths.github.io/methylit/reference/gofReport.html) search
for the best fitted model between the set of models requested by the user.

The best model including "GGamma3P" model:
```{r nlms}
d <- c("Weibull2P", "Weibull3P", "Gamma2P", "Gamma3P", "GGamma3P")
gof_1 <- gofReport(HD = HD, column = 9L, model = d, num.cores = 6L, output = "all",
                 verbose = FALSE)
gof_1$bestModel
```

The best model excluding "GGamma3P" model:
```{r nlms2}
d <- c("Weibull2P", "Weibull3P", "Gamma2P", "Gamma3P")
gof_2 <- gofReport(HD = HD, column = 9L, model = d, num.cores = 6L, output = "all",
                 verbose = FALSE)
gof_2$bestModel
```

When "GGamma3P" model is included, the goodness-of-fit indicators used here
suggest that it is best fitted model. However, a final decision must take into
account the DMP classification performance as well (see below), which is
accomplished after a signal detection step. The reader must keep in mind that we
are dealing here with numerical algorithms, which are not perfect. Depending on
dataset, the numerical computation will confront different challengers.

The difference between g3p_R.Cross.val and gg3p_R.Cross.val suggests that model
predictions powers for "GGamma3P" and "Gamma3P" are close.
```{r stats}
gof_1$stats
```

It is worthy to observe that although in the current example we successfully
performed the nonlinear fit for GGamma3P model, this is not the general case
when the number of cytosine sites goes close or above $10^6$, for which the
computational cost is very high.  The nonlinear fit of GGamma3P or GGamma4P is, in
general, difficult since two different set of estimated parameters could produce
very close distribution curves. Also, it is important to check the value of the
scaling parameter from each distribution. The numerical algorithms could find
the best fitted model with scaling parameter values close to zero, say e.g.
$10^{-8}$, and with the highest GOFs. Situations like this are biologically
meaningless and the fitted models must be discarded [@Sanchez2016].

## Signal detection
The information thermodynamics-based approach is postulated to provide greater
sensitivity for resolving true signal from the thermodynamic background within
the methylome [@Sanchez2016]. Since the biological signal created within the
dynamic methylome environment characteristic of living organisms is not free
from background noise, the approach, designated Methyl-IT, includes the
application of signal detection theory [@Greiner2000; @Carter2016; @Harpaz2013;
@Kruspe2017]. Signal detection is a critical step to increase sensitivity and
resolution of methylation signal by reducing the signal-to-noise ratio and
objectively controlling the false positive rate and prediction accuracy/risk.

## _Potential methylation signal_
The first estimation in our signal detection step is the identification of the
cytosine sites carrying potential methylation signal $PS$. The methylation
regulatory signal does not hold the theoretical distribution and, consequently, for a
given level of significance $\alpha$ (Type I error probability, e.g. 
$\alpha = 0.05$), cytosine positions $k$ with information divergence 
$H_k >= H_{\alpha = 0.05}$ can be selected as sites carrying potential signals
$PS$. The value of $\alpha$ can be specified. For example, potential signals
with $H_k > H_{\alpha = 0.01}$ can be selected. For each sample, cytosine sites
are selected based on the corresponding fitted theoretical distribution model
estimated in the previous step. Additionally, since cytosine with 
$|TV_{d_k}| < 0.1$ are the most abundant sites, depending on the sample 
(experiment), cytosine positions $k$ with $H_k >= H_{\alpha = 0.05}$ and 
$|TV_{d_k}| < 0.1$ can be observed. To prevent the last situation we can select the
$PS$ with the additional constraint $|TV_{d_k}| > TV_0$, where $TV_0$ ('tv.cut') is
a user specified value. The $PS$ is detected with the function 
[getPotentialDIMP](https://genomaths.github.io/methylit/reference/getPotentialDIMP.html):

```{r getPotentialDIMP}
PS1 = getPotentialDIMP(LR = HD, dist.name = gof_1$bestModel, nlms = gof_1$nlms,
                      div.col = 9, alpha = 0.05, tv.col = 8, tv.cut =  0.5)
PS2 = getPotentialDIMP(LR = HD, dist.name = gof_2$bestModel, nlms = gof_2$nlms,
                      div.col = 9, alpha = 0.05, tv.col = 8, tv.cut =  0.5)
PS1$Breast_normal
PS2$Breast_normal
```

Model "GGamma3P" reported 20263 - 11314 = 8949 more DMPs in the normal tissue
than model "Gamma3P".

Notice that the total variation distance $TV_d=|TV|$ is an information
divergence as well and that it can be used in place of Hellinger divergence
[@Sanchez2016]. The set of vectors $P_i = (p_i, 1-p_i)$ and distance function
$TV_d$ integrate a metric space [@Sanchez2019]. In particular, after takes the Manhattan
distance between healthy individuals and patients methylation levels we have:
$$M_d(p_{ij}, p_{ir})=\frac{1}{2}(|\hat p_{ij} - \hat p_{ir}| + |(1-\hat p_{ij}) - (1-\hat
p_{ir})|)=|\hat p_{ij} - \hat p_{ir}|=TV_d(p_{ij}, p_{ir})$$. 

That is, the absolute difference of methylation levels is a particular case of
Manhattan distance, which is named total variation distance $TV_d$. The
quantitative effect of the vector components $1-\hat p_{ij}$ and $1-\hat p_{ir}$
(in our case, the effect of unmethylated read counts) is not present in $TV_d$ as
in $H(\hat p_{ij},\hat p_{ir})$.

## _Histogram and boxplots of methylation potential signals_
As before, a single GRanges object is built from the above set GRanges objects
using the function [uniqueGRanges](https://genomaths.github.io/methylit/reference/uniqueGRanges.html),
and the Hellinger divergences of the cytosine sites carrying $PS$ (for each
sample) are located in a single matrix on the metacolumn of the GRanges object.

```{r fig2, fig.height = 14, fig.width = 20}
# Define an auxiliar function
fun_length <- function(x){
  return(data.frame(y = median(x) + 1.5, label = paste0("n = ", length(x))))
}

# The data to be used in the boxplot
ps1 = uniqueGRanges(PS1, missing = NA, columns = 9, verbose = FALSE, num.cores = 6L)
colnames(mcols(ps1)) <- c("normal", "cancer", "metastasis", "normal1")
dat1 = data.frame(normal = ps1$normal, normal1 = ps1$normal1, cancer = ps1$cancer,  
                 metastasis = ps1$metastasis)
dat1 = suppressMessages(melt(dat1))
colnames(dat1) <- c("Breast.tissue", "HD")
idx <- which(is.na(dat1$HD))
dat1 <- dat1[-idx, ] # To remove missing data

ps2 <- uniqueGRanges(PS2, missing = NA, columns = 9, verbose = FALSE, num.cores = 6L)
colnames(mcols(ps2)) <- c("normal", "cancer", "metastasis", "normal1")
dat2 = data.frame(normal = ps2$normal, normal1 = ps2$normal1, cancer = ps2$cancer,  
                 metastasis = ps2$metastasis)
dat2 = suppressMessages(melt(dat2))
colnames(dat2) <- c("Breast.tissue", "HD")
idx <- which(is.na(dat2$HD))
dat2 <- dat2[-idx, ] # To remove missing data

p1 = ggplot(dat1, aes(x = HD, fill = Breast.tissue, colour = Breast.tissue)) + 
  geom_histogram(alpha = 0.5, bins = 50, position = "identity", na.rm = TRUE,
                 size = 0.7)  + ylab("Counts") + xlim(0,30) +
theme(axis.title.x = element_text(face = "bold", size = 20),
        axis.text.x = element_text(face = "bold", size = 20, color = "black",
                                   hjust = 0.5, vjust = 0.75),
        axis.text.y = element_text(face = "bold", size = 20, color = "black"),
        axis.title.y = element_text(face = "bold", size = 20,color = "black"),
        legend.text = element_text(size = 20, face = "bold"),
        legend.title = element_text(size = 20, face = "bold")
        ) +
    ggtitle("Histogram for Potential methylation signal. GGamma3P")

p2 = ggplot(dat1,aes(x = Breast.tissue, y = HD , fill = Breast.tissue)) +
  geom_boxplot(na.rm = TRUE) + ylim(0, 15) +
  stat_summary(fun.data = fun_length, geom = "text", na.rm = TRUE, 
               position = position_dodge(width = 0.9), vjust = 1,
               size = 6, fontface = "bold") +
  geom_hline(yintercept = 6, linetype="dashed", color = "red") +
  theme(axis.title.x = element_text(face = "bold", size = 20),
        axis.text.x = element_text(face = "bold", size = 20, color = "black",
                                   hjust = 0.5, vjust = 0.75),
        axis.text.y = element_text(face = "bold", size = 20, color = "black"),
        axis.title.y = element_text(face = "bold", size = 20,color = "black"),
        legend.position = "none"
        ) +
      ggtitle("Boxplot for Potential methylation signal. GGamma3P")


p3 = ggplot(dat2, aes(x = HD, fill = Breast.tissue, colour = Breast.tissue)) + 
  geom_histogram(alpha = 0.5, bins = 50, position = "identity", na.rm = TRUE,
                 size = 0.7)  + ylab("Counts") + xlim(0,30) +
theme(axis.title.x = element_text(face = "bold", size = 20),
        axis.text.x = element_text(face = "bold", size = 20, color = "black",
                                   hjust = 0.5, vjust = 0.75),
        axis.text.y = element_text(face = "bold", size = 20, color = "black"),
        axis.title.y = element_text(face = "bold", size = 20,color = "black"),
        legend.text = element_text(size = 20, face = "bold"),
        legend.title = element_text(size = 20, face = "bold")
        ) +
    ggtitle("Histogram for Potential methylation signal. Gamma3P")

p4 = ggplot(dat2,aes(x = Breast.tissue, y = HD , fill = Breast.tissue)) +
  geom_boxplot(na.rm = TRUE) + ylim(0, 15) +
  stat_summary(fun.data = fun_length, geom = "text", na.rm = TRUE, 
               position = position_dodge(width = 0.9), vjust = 1,
               size = 6, fontface = "bold") +
  geom_hline(yintercept = 6, linetype="dashed", color = "red") +
  theme(axis.title.x = element_text(face = "bold", size = 20),
        axis.text.x = element_text(face = "bold", size = 20, color = "black",
                                   hjust = 0.5, vjust = 0.75),
        axis.text.y = element_text(face = "bold", size = 20, color = "black"),
        axis.title.y = element_text(face = "bold", size = 20,color = "black"),
        legend.position = "none"
        ) +
      ggtitle("Boxplot for Potential methylation signal. Gamma3P")


grid.arrange(p1, p2, p3, p4, ncol = 2)

```

The graphics show that "Gamma3P" provides more information on the discrimination
of treatment potential DMPs from the control.

## Cutpoint estimation
Laws of statistical physics can account for background methylation, a response
to thermal fluctuations that presumably functions in DNA stability
[@Sanchez2016; @Sanchez2019]. True signal is detected based on the optimal
cutpoint [@lopez2014], which can be estimated from the area under the curve
(AUC) of a receiver operating characteristic (ROC) curve built from a logistic
regression performed with the potential signals from controls and treatments.
The ROC AUC is equivalent to the probability that a randomly-chosen positive
instance is ranked more highly than a randomly-chosen negative instance
[@Fawcett2005]. In the current context, the AUC is equivalent to the probability
to distinguish a randomly-chosen methylation regulatory signal induced by the
treatment from a randomly-chosen signal in the control.

The need for the application of (what is now known as) signal detection in
cancer research was pointed out by Youden in the midst of the last century
[@Youden1950]. In the next example, the simple cutpoint estimation available in
Methyl-IT is based on the application of [Youden
index](https://en.wikipedia.org/wiki/Youden%27s_J_statistic) [@Youden1950].
Although cutpoints are estimated for a single variable, the classification
performance can be evaluated for several variables and applying different model
classifiers. A optimal cutpoint distinguishes disease stages from healthy
individual. The performance of this classification is given in the output of
function
[estimateCutPoint](https://genomaths.github.io/methylit/reference/estimateCutPoint.html).
A model classifier can be requested for further predictions and its
classification performance is also provided. Below, the selected model
classifier is a quadratic discriminant analysis (QDA) (*classifier1* = "qda",
*clas.perf* = TRUE). Four predictor variables are available: the Hellinger
divergence of methylation levels (*hdiv*), total variation distance (*TV*,
absolute difference of methylation levels), relative position of cytosine site
in the chromosome (*pos*), and the logarithm base two of the probability to
observe a Hellinger divergence value $H$ greater than the critical value
$H_{\alpha = 0.05}$ (values given as probabilities in object PS, *wprob*). 

Notice that the cutpoint can be estimated for any of the two currently available
information divergences: *Hellinger divergence* (*div.col* = 9) or the *total*
*variation distance* (with Bayesian correction, *div.col* = 8).

### _Cutpoint based on *Youden index*_

Null distribution "GGamma3P":
```{r cutpoints}
cutpoint1 = estimateCutPoint(LR = PS1, 
                            control.names = c("Breast_normal","Breast_normal1"), 
                            treatment.names = c("Breast_cancer", "Breast_metastasis"),
                            simple = TRUE,
                            classifier1 = "pca.logistic", 
                            column = c(hdiv = TRUE, bay.TV = TRUE, 
                                       wprob = TRUE, pos = TRUE),
                            n.pc = 4 , center = TRUE, scale = TRUE,
                            div.col = 9, clas.perf = TRUE)
cutpoint1$cutpoint
cutpoint1$testSetPerformance
cutpoint1$testSetModel.FDR
```

Null distribution "Gamma3P":
```{r cutpoints2}
cutpoint2 = estimateCutPoint(LR = PS2, 
                            control.names = c("Breast_normal","Breast_normal1"), 
                            treatment.names = c("Breast_cancer", "Breast_metastasis"),
                            simple = TRUE,
                            classifier1 = "pca.logistic", 
                            column = c(hdiv = TRUE, bay.TV = TRUE, 
                                       wprob = TRUE, pos = TRUE),
                            n.pc = 4 , center = TRUE, scale = TRUE,
                            div.col = 9, clas.perf = TRUE)
cutpoint2$cutpoint
cutpoint2$testSetPerformance
cutpoint2$testSetModel.FDR

```

The cutpoint estimated based “GGamma3P” model is lower than the value based on
“Gamma3P” model. Often, this is a good detail, since it permits the
identification of more sites with high probability to carry methylation signal.
However, the evaluation of the classification performance has the last word on
model selection. In the current case, the DMP classification performance is much
better for “Gamma3P” model than for “GGamma3P”.  In particular, the
probabilities derived from each distribution model provide fundamental
information to the model classifier. Hence, the probability that a potential DMP
could carry a methylation signal estimated using "Gamma3P" model is more
informative (for the model classifier) than the probability estimated using
"GGamma3P" model.

### _Cutpoint estimated using a model classifier_
In the current case, the DMP classification performance for the signal derived 
with null distribution "Gamma3P" can be improved searching for an optimal 
cutpoint with a model classifier and using the total variation distance (
with Bayesian correction, bay.TV).

```{r cutpoints3}
cutpoint2 = estimateCutPoint(LR = PS2, 
                            control.names = c("Breast_normal","Breast_normal1"), 
                            treatment.names = c("Breast_cancer", "Breast_metastasis"),
                            simple = FALSE,
                            classifier1 = "pca.logistic", 
                            classifier2 = "pca.qda",
                            column = c(hdiv = TRUE, bay.TV = TRUE, 
                                       wprob = TRUE, pos = TRUE),
                            n.pc = 4 , center = TRUE, scale = TRUE,
                            div.col = 8)
cutpoint2$cutpoint
cutpoint2$testSetPerformance
cutpoint2$testSetModel.FDR
```

The good DMP classification performance was kept at lower cutpoint value.

## DMPs
Cytosine sites carrying a methylation signal are designated differentially
informative methylated positions (DMPs). The probability that a DMP is not
induced by the treatment is given by the probability of false alarm (P_FA, false
positive, Sanchez et al. 2019). That is, the biological signal is naturally
present in the control as well as in the treatment. According to the classical
view in the epigenomic field, a DMP is a cytosine position carrying a
significant methylation signal, which might be identified as differentially
methylated cytosine site by Fisher’s exact test (or other current tests). _*In
Methyl-IT, a DMP is a DNA cytosine position with high probability to be
differentially hyper-methylated or hypo-methylated in the treatment with respect
to a given control (reference individual)*_. Notice that Methyl-IT definition of
DMP is not deterministic in an ordinary sense, but stochastic-deterministic in
physico-mathematical terms. In other words, if a given cytosine position is
identified as a hyper-methylated DMP in the treatment, then with high
probability such a site will be found more frequently hyper-methylated than
hypo-methylated in the _*treatment population*_. Observe that it does not mean
that, for all arbitrary subset of samples, the assumed cytosine position will be
hypermethylated.

DMPs are selected with the function [selectDIMP](https://genomaths.github.io/methylit/reference/selectDIMP.html)
```{r DMPs}
DMPs = selectDIMP(PS2, div.col = 8, cutpoint = cutpoint2$cutpoint)

```

### The descriptive statistic of the read counts at DMPs

After the signal detection step the read counts at DMPs is notable greater than
at the general methylated cytosine background:
```{r meth-counts}

# Maximum coverage at each cytosine site
methc <- lapply(DMPs, function(x) {
     r <- apply(cbind(x$c1, x$c2), 1, max)
     return(r)
})

# Quantiles
do.call(rbind, lapply(methc, function(x) {
    q10 <- quantile(x, 0.1)
    q60 <- quantile(x, 0.6)
    q9999 <- quantile(x, 0.9999)
    idx1 <- which(x >= q60)
    idx2 <- which(x <= 500)
    q95 <- quantile(x, 0.95)

    idx <- intersect(idx1, idx2)
    return(c(round(summary(x)), q10, q60, 
             quantile(x, c(0.95, 0.99, 0.999, 0.9999)), 
             num.siteGreater_8 = sum(x >= 8), 
             q60_to_500 = sum((x >= q60) & (x <= 500)),
             num.siteGreater_500 = sum(x > 500)))})
)
```

The minimum amount of methylated read counts is observed "Breast_metastasis" (15 reads)
and for any the samples the tenth percent of DMPs is 26 or more reads.

### _Venn Diagram of DMPs_
The Venn diagram of DMPs reveals that the number cytosine site carrying
methylation signal with a divergence level comparable to that observed in
breast tissues with cancer and metastasis is relatively small (2797 DMPs). The
number of DMPs decreased in the breast tissue with metastasis, but, as shown
in the last boxplot, the intensity of the signal increased.

```{r fig4, fig.height = 5, fig.width = 7}
n12 = length(GenomicRanges::intersect(DMPs$Breast_normal,
                                      DMPs$Breast_cancer))
n13 = length(GenomicRanges::intersect(DMPs$Breast_normal,
                                      DMPs$Breast_metastasis))
n23 = length(GenomicRanges::intersect(DMPs$Breast_cancer,
                                      DMPs$Breast_metastasis))
n123 = length(Reduce(GenomicRanges::intersect, 
                     list(DMPs$Breast_normal, DMPs$Breast_cancer,
                          DMPs$Breast_metastasis)))

grid.newpage()
v = draw.triple.venn(area1 = length(DMPs$Breast_normal),
                     area2 = length(DMPs$Breast_cancer),
                     area3 = length(DMPs$Breast_metastasis),
                     n12 = n12, n23 = n23, n13 = n13, n123 = n123,
                     category = c("Breast_normal", "Breast_cancer", 
                                  "Breast_metastasis"),
                     lty = rep("blank", 3), fill = c("blue", "yellow",
                                                     "magenta"),
                     alpha = c(0.1, 0.2, 0.3), 
                     cat.pos = c(-80, 90, 0),
                     cat.col = c("blue", "darkorange", "red"),
                     cat.dist = c( -0.1, -0.08, -0.26),
                     cex = rep(1.7, 7),
                     cat.cex = c( 1.5, 1.5, 1.5),
                     label.col = c( "blue", "darkorange", "darkorange",
                                    "red", 
                                    "white", "red", "red"),
                     scaled = TRUE)
grid.draw(v)


```


Notice that natural methylation regulatory signals (not induced by the
treatment) are present in both groups, control and treatment. The signal
detection step permits us to discriminate the "ordinary" signals observed in
the control from those induced by the treatment (a disease in the current
case). In addition, this diagram reflects a classification of DMPs only based
on the cytosine positions. That is, this Venn diagram cannot tell us whether
DMPs at the same position can be distinguishable or not. For example, DMPs at
the same positions in control and treatment can happen with different
probabilities estimated from their corresponding fitted Weibull distributions
(see below).

### Monte Carlo Evaluation of the Model Classification Performance
The regulatory methylation signal is an output from a natural process that
continuously takes place across the ontogenetic development of the organism.
Therefore, we expect to see methylation signal in natural, ordinary conditions.
Function 'evaluateDIMPclass' can be used to perform a classification of DMPs
into two classes: DMPS from control and DMPs from treatment samples, as well
as an evaluation of the classification performance (for more details see
[evaluateDIMPclass](https://genomaths.github.io/methylit/reference/evaluateDIMPclass.html)). 
As a matter of fact this function is called by function [estimateCutPoint](https://genomaths.github.io/methylit/reference/estimateCutPoint.html).
Additional feature not used before is the possibility to perform Monte Carlo evaluation 
of the model classifier performance. In above split of the sample into two subsets, 
training and test datasets, we would be just lucky getting the right proportion 
that yields the best classification performance. What about is the random split is
repeat 300 times?

The performance of the "pca.qda" classifier is:
```{r mc1}
performance <- evaluateDIMPclass(LR = DMPs, 
                                 control.names = c("Breast_normal", "Breast_normal1"),
                                 treatment.names = c("Breast_cancer", "Breast_metastasis"),
                                 column = c(hdiv = TRUE, bay.TV = TRUE, 
                                            wprob = TRUE, pos = TRUE),
                                 classifier = "pca.qda", prop = 0.6, 
                                 n.pc = 4 , center = TRUE, scale = TRUE,
                                 num.boot = 300, output = "mc.val", 
                                 num.cores = 6L, tasks = 2L)
performance
```


## Differentially methylated genomic regions (DMRs)
Our degree of confidence in whether DMP counts in both groups of samples,
control and treatment, represent true biological signal was determined in the
signal detection step. To estimate DIMRs, we followed similar steps to those
proposed in Bioconductor R package DESeq2 [@Love2014], but our GLM test looks
for statistical difference between the groups based on gene-body DMP counts
overlapping a given genomic region rather than read counts. The regression
analysis of the generalized linear model (GLM) with logarithmic link was
applied to test the difference between group counts. The fitting algorithmic
approaches provided by 'glm' and 'glm.nb' functions from the R packages stat
and MASS, respectively, were used for Poisson (PR), Quasi-Poisson (QPR) and
Negative Binomial (NBR) linear regression analyses, respectively.

In Methyl-IT, the concept of DMR is generalized and it is not limited to any
particular genomic region found with specific clustering algorithm. It can be
applied to any naturally or algorithmically defined genomic region. For example,
an exon region identified statistically to be differentially methylated by using
generalized linear regression model (GLM) is a DMR. Differentially methylated
genes (DMGs) are estimated from group comparisons for the number of DMPs on
gene-body regions between control and treatment.  This permits the extension of
the concept of DMR by considering a differentially methylated gene as a
particular case of DMRs.

### _Differentially methylated genes (DMGs)_
We shall call DMGs those DIMRs restricted to gene-body regions. DMGs are
detected using function [countTest2](https://genomaths.github.io/methylit/reference/countTest2.html).

Gene annotation is taken from [Ensembl](https://www.ensembl.org/) 
```{r AG, eval = FALSE}
# To load human gene annotation 
# AG = import(con = paste0("ftp://ftp.ensembl.org/pub/release-91/gff3/",
#                          "homo_sapiens/Homo_sapiens.GRCh38.91.gff3.gz"))
# genes = AG[ AG$type == "gene", c( "gene_id", "biotype", "Name" ) ]
# genes = genes[ genes$biotype == "protein_coding", "gene_id" ]
# seqlevels(genes, "coarse") <- "13" # To keep a consistent chromosome annotation
# seqlevels(genes) <- "chr13"

```

Here, for the sake of brevity, we just load the "RData" compressed file 
containing only the gene annotation:
```{r sake}
con = url("https://git.psu.edu/genomath/MethylIT_examples/raw/master/Homo_sapiens_GRCh38.91_genes.RData")
load(con)
genes = genes[ genes$biotype == "protein_coding", ]
seqlevels(genes, "coarse") <- "13" # To keep a consistent chromosome annotation
seqlevels(genes) <- "chr13"
genes
```

Function [getDIMPatGenes](https://genomaths.github.io/methylit/reference/getDIMPatGenes.html)
is used to count the number of DMPs at gene-body. Nevertheless, it can be used
for any arbitrary specified genomic region as well. The operation of this
function is based on the '*findOverlaps*' function from the '*GenomicRanges*'
Bioconductor R package. The '*findOverlaps*' function has several critical
parameters like, for example, '*maxgap*', '*minoverlap*', and '*ignore.strand*'.
In our function
[getDIMPatGenes](https://genomaths.github.io/methylit/reference/getDIMPatGenes.html),
except for setting ignore.strand = TRUE and *type = "within"*, we preserve the rest of default
'findOverlaps' parameters. In this case, these are important parameter settings
because the local mechanical effect of methylation changes on a DNA region
where a gene is located is independent of the strand where the gene is encoded.
That is, methylation changes located in any of the two DNA strands inside the
gene-body region will affect the flexibility of the DNA molecule [@Choy2010;
@Severin2011].

```{r getDIMPatGenes}
DMPsBN = getDIMPatGenes(GR = DMPs$Breast_normal, GENES = genes)
DMPsBN1 = getDIMPatGenes(GR = DMPs$Breast_normal1, GENES = genes)
DMPsBC = getDIMPatGenes(GR = DMPs$Breast_cancer, GENES = genes)
DMPsBM = getDIMPatGenes(GR = DMPs$Breast_metastasis, GENES = genes)
```

The number of DMPs on the strand where a gene is encoded is obtained by
setting ignore.strand = FALSE. However, for the current example results will be
the same since the datasets downloaded from GEO do not have strand information.
Next, the above GRanges objects carrying the DMP counts from each sample are
grouped into a single GRanges object. Since we have only one control, to
perform group comparison and to move forward with this example, we duplicated
'Breast_normal' sample. Obviously, the confidence on the results increases with
the number of sample replications per group (in this case, it is only an
illustrative example on how to perform the analysis, since a fair comparison
requires for more than one replicate in the control group).

```{r uniqueGRanges}
Genes.DMPs = uniqueGRanges( list(DMPsBN[, 2], DMPsBN1[, 2],
                                  DMPsBC[, 2], DMPsBM[, 2]),
                             type = "equal", verbose = FALSE,
                             ignore.strand = TRUE )
colnames( mcols(Genes.DMPs)) <- c("Breast_normal", "Breast_normal1",
                                   "Breast_cancer", "Breast_metastasis")

```

Next, the set of mapped genes are annotated 
```{r subsetByOverlaps}
GeneID = subsetByOverlaps(genes, Genes.DMPs, type = "equal", 
                          ignore.strand = FALSE)
names( Genes.DMPs ) <- GeneID$gene_id
Genes.DMPs
```

Now, we build a [glmDataSet](https://genomaths.github.io/methylit/reference/glmDataSet.html) 
object.  
```{r glmDataSet}
## An experiment design is set.
colData <- data.frame(condition = factor(c("BN","BN","BC","BC"),
                                         levels = c("BN", "BC")),
                       c("Breast_normal","Breast_normal1",
                         "Breast_cancer","Breast_metastasis"),
                       row.names = 2)
## A RangedGlmDataSet is created
ds <- glmDataSet(GR = Genes.DMPs, colData = colData)
```

DMG analysis is performed with the function [countTest2](https://genomaths.github.io/methylit/reference/countTest2.html)
```{r countTest}
DMGs = countTest2(DS = ds, num.cores = 4L, minCountPerIndv = 8, 
                  countFilter = TRUE, maxGrpCV = c(1, 1),
                 Minlog2FC = 1, pvalCutOff = 0.05,
                 MVrate = .95, verbose = FALSE)
DMGs
```

Here, we adds the gene alias:
```{r alias}
hits = findOverlaps(DMGs, genes, type = "equal")
DMGs = DMGs[queryHits(hits)]
DMGs$alias <- genes$Name[subjectHits(hits)]
```

_**BRCA2, a breast cancer associated risk gene, is found between the DMG**s_
```{r DMGs0}
# DMGs
DMGs[ grep( "ENSG00000139618", names(DMGs) ) ]
```

## Concluding remarks

Results fundamentally depend on the use of the proper reference samples. If there is
not enough control samples to create an independent reference virtual individual, then
the centroid from the control group can be used instead, which can be built with function 
[poolFromGRlist](https://genomaths.github.io/methylit/reference/poolFromGRlist.html).

The *reference* is used to put individuals from control and patients in the same
*system of reference*. The information divergences of methylation levels with
respect to the reference individual are estimated for each cytosine site for
each individual, from control and treatment. As a result, the information on the
natural spontaneous variation in the control and treatment populations is
carried in the probability distributions of the information divergences,
estimated for each individual. *It does not matter how statistically*
*significant a DMP would be in the group of patients*. *What really matter is
how big is the* *probability to observe the same methylation event in the control
group in respect to the patient group*. Decisions are ultimately made on the
basis of such classification probabilities at single cytosine positions. The
signal detection based approach implemented in Methyl-IT is addressed to confront
the mentioned issue and, in consequence, the level of resolution of Methyl-IT
approach reach the single cytosine site.


## Acknowledgments
We thank Professor David J Miller for valuable conversations and suggestions on
our mathematical modeling.
## Funding The work was supported by funding from NSF-SBIR
(2015-33610-23428-UNL) and the Bill and Melinda Gates Foundation (OPP1088661).

## S2. Session Information

```{r, echo=FALSE}
sessionInfo()

```


## References. 

