---
title: "DMR detection with Methyl-IT"
subtitle: "Analysis for Arabidopsis MSH1-memory versus non-memory first generation"
author: |
     | Robersy Sanchez
     | rus547@psu.edu
     | Mackenzie's lab
     |
     | Department of Biology and Plant Science. 
     | Pennsylvania State University, University Park, PA 16802
     | Maintainer: Robersy Sanchez
date: "`r format(Sys.time(), '%d %B %Y')`"
fontsize: 11pt
fontfamily: "serif"
bibliography: bibliography.bib
output:
    rmarkdown::html_vignette:
        toc: true
        toc_depth: 4
        number_sections: true
        geometry: margin=1in
        keep_md: yes
vignette: >
    %\VignetteIndexEntry{DMR detection with Methyl-IT}
    %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
abstract: |
    Herein, an example of Methyl-IT application to the detection of differential
    methylated regions (DMRs) on first generation MSH1 RNAi transgene segregants
    data set is presented.  
---


<style type="text/css">

.list-group-item.active, .list-group-item.active:focus,
.list-group-item.active:hover {
    z-index: 2;
    color: #fff;
    background-color: #337ab7;
    border-color: #337ab7;
}

.tocify-subheader .tocify-item {
    font-size: 0.80em;
    padding-left: 25px;
    text-indent: 0;
}

.tocify-header .tocify-item {
    font-size: 0.85em;
    padding-left: 20px;
    text-indent: 0;
}

body{ /* Normal */
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    text-align: justify
}
td { /* Table */
    font-size: 8px;
}

h1.title {
    font-size: 38px;
    font-family: "Times New Roman", Times, serif;
    color: DarkRed;
    .subTitle {
        font-size: 24px;
        font-family: "Times New Roman", Times, serif;
        color: DarkRed;
    }
}

h1 { /* Header 1 */
    font-size: 28px;
    font-family: "Times New Roman", Times, serif;
    color: DarkBlue;
}
h2 { /* Header 2 */
    font-size: 22px;
    color: DarkBlue;
    font-family: "Times New Roman", Times, serif;
}
h3 { /* Header 3 */
    font-size: 18px;
    color: DarkBlue;
    font-family: "Times New Roman", Times, serif;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(options(width = 80))
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Introduction  
MethylIT is an R package for methylome analysis based on a signal-detection
machine learning approach (SD-ML). This approach is postulated to provide
greater sensitivity for resolving true signal from the methylation background
within the methylome [@Sanchez2016; @Sanchez2019]. Because the biological signal
created within the dynamic methylome environment characteristic of plants is not
free from background noise, the approach, designated Methyl-IT, includes the
application of signal detection theory [@Greiner2000;
@Carter2016; @Harpaz2013; @Kruspe2017].

A basic requirement for the application of signal detection is the
knowledge of the probability distribution of the background noise, which is used
as null hypothesis in the detection of the methylation signal. This probability
distribution is, in general, a member of generalized gamma distribution family,
and it can be deduced on a statistical mechanical/thermodynamics basis from DNA
methylation induced by thermal fluctuations [@Sanchez2016].

Herein, we provide an example of Methyl-IT application to DMR estimation on
patients with PALL [@Wahlberg2016]. Due to the size of human methylome the
current example only covers the analysis of chromosome 9. A full description of
Methyl-IT application of methylome analysis for Clinical Diagnostics is given in
the manuscript [@Sanchez2019]. A wider analysis of PALL data set is given in
[@Sanchez2020].

The first step of DMR estimation follows the general step of Methyl-IT pipeline
for the estimation of DMPs. In Methyl-IT, a DMP is a cytosine positions that,
with high probability, carries a statistically significant methylation change in
respect to the same position in a control population. Once DMPs are identified, 
a simple algorithm can be applied to identify clusters of DMPs using [dmpClusters](https://genomaths.github.io/methylit/reference/dmpClusters.html)
function. Next, [countTest2](https://is.gd/kltCXl) is applied to test whether a
given cluster of DMPs is a DMR.

It is worthy to notices that any clustering algorithm will impose non-natural
constraints on the methylation patterning, which not necessarily will be
associated to the local DNA physico-chemical properties and to local molecular
evolutionary features of every DMR detected. Every cytosine methylation has a
concrete physical effect of the DNA, altering the DNA persistence length, i.e.,
altering the local DNA flexibility [@Severin2011; @Perez2012; @Yusufaly2013].
DNA flexibility was shown to affect the binding of proteins to methylated DNA
and DNA-containing lesion [@Teng2018; @Hodges2020] and nucleosome mechanical
stability [@Ngo2016]. In other words, the sizes of naturally occurrying
methylation motifs are not arbitrarily set, but very well constrained
under molecular evolutionary pressure, probably, particular for each species.

The R packages required in this example are:
```{r libraries, message=FALSE, results='hide'}
        library(MethylIT)
        library(MethylIT.utils)
        library(ggplot2) # graphic
        library(grid) # For multiple plots
        library(gridExtra) # For multiple plots
        library(rtracklayer) # To import gene annotation
```


# Estimation of differentially methylated regions (DMRs)

DMP data set is available for download.  
```{r dmps, message=FALSE, results='hide'}
## ===== Download methylation data from Github ========
url <- paste0("https://github.com/genomaths/genomaths.github.io/raw/master/AT_memory/", 
              "dmps_mm-vs-nm_all-context_gen1_nmRef_min-meth-3_cov-4_04-16-20.RData")
temp <- tempfile(fileext = ".RData")
download.file(url = url, destfile = temp)
load(temp)
file.remove(temp); rm(temp, url)

```

The gene annotation
```{r annotation, message=FALSE, results='hide'}
## ===== Download methylation data from GitHub ========
url <- paste0("https://github.com/genomaths/genomaths.github.io/raw/master/", 
              "Annotation/Arabidopsis_thaliana_genes_TAIR10.38.gtf.RData")
temp <- tempfile(fileext = ".RData")
download.file(url = url, destfile = temp)
load(temp)
file.remove(temp); rm(temp, url);

GENES <- genes[genes$type == "gene"]

## To get exons and introns
exons <- genes[genes$type == "exon"]
introns <- setdiff(x = genes, y = exons, ignore.strand = TRUE)

## To add the corresponding gene names to introns
hits <- findOverlaps(introns, genes, type = "within", ignore.strand = TRUE)
introns <- introns[queryHits(hits)]
mcols(introns) <- mcols(genes[subjectHits(hits)])

```

## Cluster of DMPs

Function [dmpClusters](https://is.gd/K2cRgZ) search for cluster of DMPs found
from all the samples. The basic idea reside in that if a given cluster of DMP is
a DMR, then it will be consistently differentially methylated in all treatment
samples, and we can identify them applying function [countTest2](https://
is.gd/kltCXl).

There is not a unique approach to build DMP clusters, since several clustering
criteria can be applied. Herein, we use the two approach implemented in  [dmpClusters](https://is.gd/K2cRgZ) function (follow the link or type 
?dmpClusters to see the help for the function).

```{r clusters_fixed.int}
clust_50_100 = dmpClusters(GR = dmps, maxDist = 50, minNumDMPs = 10, 
                           maxClustDist = 100, method = "fixed.int",
                           num.cores = 1L, verbose = FALSE)

clust_50_100
```

DMPs clusterization strongly depends on the maximum distance permitted between 
two consecutive DMPs that belong to the same cluster (parameter: `minNumDMPs`) 
and on the minimum number of DMPs that must be present on a detected 
agglomeration of DMPs to be considered a cluster (parameter: `maxDist`).
```{r clusters_relaxed}
.clust_50_100 = dmpClusters(GR = dmps, maxDist = 50, minNumDMPs = 10, 
                           maxClustDist = 100, method = "relaxed",
                           num.cores = 1L, verbose = FALSE)
.clust_50_100
```


Alternatively, we can use function _dmpClustering_ from MethylIT.utils R package.
Notice that although _'method'_ parameter are named the same, on each function
these stand for different approaches.
```{r clusters.fixed.int}
clust.50.100 = dmpClustering(dmps = dmps, minNumDMPs = 10, win.size = 50,
                            step.size = 50, method = "fixed.int", 
                            maxClustDist = 100, verbose = FALSE)
clust.50.100
```


```{r clusters.relaxed}
.clust.50.100 = dmpClustering(dmps = dmps, minNumDMPs = 10, method = "relaxed", 
                            maxClustDist = 100, verbose = FALSE)
.clust.50.100
```


## DMR detection

The count of the number of DMPs on each region will be accomplished with
[getDMPatRegions](https://genomaths.github.io/methylit/reference/getDMPatRegions.html)
function. This function operates based on [*findOverlaps*](https://is.gd/zJ7U9l)
function from the '*GenomicRanges*' Bioconductor R package. The
[*findOverlaps*](https://is.gd/zJ7U9l) function has several critical parameters
like, for example, '*maxgap*', '*minoverlap*', and '*ignore.strand*'. In our
[getDMPatRegions](https://genomaths.github.io/methylit/reference/getDMPatRegions.html)
function, except for setting ignore.strand = TRUE and *type = "within"*, we
preserve the rest of default 'findOverlaps' parameters.

### Experiment design for each clustering approach 

Design for method = "fixed.int" from dmpClusters function 
```{r design_fixed.int}
## ==== Setting up the experiment design to test for DMRs ===
nams <- names(dmps)
dmps_at_clust_50_100 <- getDMPatRegions(GR = dmps, regions = clust_50_100,
                                       ignore.strand = TRUE)
dmps_at_clust_50_100 <- uniqueGRanges(dmps_at_clust_50_100, columns = 2L,
                                      ignore.strand = TRUE, type = 'equal',
                                      verbose = FALSE)
colnames(mcols(dmps_at_clust_50_100)) <- nams

colData <- data.frame(condition = factor(c("MM", "MM","MM","MM", "MM", 
                                           "NM","NM","NM","NM"),
                                         levels = c("NM", "MM")),
                      nams,
                      row.names = 2)

## Build a RangedGlmDataSet object
ds_50_100 <- glmDataSet(GR = dmps_at_clust_50_100, colData = colData)

```

The GRanges object with DMPs on it
```{r dmr_dat}
dmps_at_clust_50_100
```

The [glmDataSet](https://is.gd/VZaazz) object.
```{r glm}
ds_50_100
```


Design for method = "relaxed" from dmpClusters function 
```{r design_relaxed}
## ==== Setting up the experiment design to test for DMRs ===
nams <- names(dmps)
.dmps_at_clust_50_100 <- getDMPatRegions(GR = dmps, regions = .clust_50_100,
                                       ignore.strand = TRUE)
.dmps_at_clust_50_100 <- uniqueGRanges(.dmps_at_clust_50_100, columns = 2L,
                                      ignore.strand = TRUE, type = 'equal',
                                      verbose = FALSE)
colnames(mcols(.dmps_at_clust_50_100)) <- nams

## Build a RangedGlmDataSet object
.ds_50_100 <- glmDataSet(GR = .dmps_at_clust_50_100, colData = colData)

```

Design for method = "fixed.int" from dmpClustering function 
```{r design.fixed.int}
## ==== Setting up the experiment design to test for DMRs ===
nams <- names(dmps)
dmps_at_clust.50.100 <- getDMPatRegions(GR = dmps, regions = clust.50.100,
                                       ignore.strand = TRUE)
dmps_at_clust.50.100 <- uniqueGRanges(dmps_at_clust.50.100, columns = 2L,
                                      ignore.strand = TRUE, type = 'equal',
                                      verbose = FALSE)
colnames(mcols(dmps_at_clust.50.100)) <- nams

## Build a RangedGlmDataSet object
ds.50.100 <- glmDataSet(GR = dmps_at_clust.50.100, colData = colData)

```


Design for method = "relaxed" from dmpClustering function 
```{r design.relaxed}
## ==== Setting up the experiment design to test for DMRs ===
nams <- names(dmps)
dmps_at.clust.50.100 <- getDMPatRegions(GR = dmps, regions = .clust.50.100,
                                       ignore.strand = TRUE)
dmps_at.clust.50.100 <- uniqueGRanges(dmps_at.clust.50.100, columns = 2L,
                                      ignore.strand = TRUE, type = 'equal',
                                      verbose = FALSE)
colnames(mcols(dmps_at.clust.50.100)) <- nams

## Build a RangedGlmDataSet object
ds_50.100 <- glmDataSet(GR = dmps_at.clust.50.100, colData = colData)

```


### DMR detection

Herein, we are requesting a minimum density DMP of 1/100 bp = 0.01/bp (CountPerBp = 0.01), minimum 
of logarithm of fold changes in the number of DMPs, treatment versus control 
samples, equal to 1 (Minlog2FC = 1). Also, it is requested a coefficient of 
variation, on each sample set, not greater than 1 (maxGrpCV = c(1, 1)). 

Notice that in the case of species with big genes, for example, human genes, we
might want to use lower densities, e.g., CountPerBp = 0.001.

DMRs derived method = "fixed.int" from dmpClusters function 
```{r dmrs_100, message=FALSE, results='hide'}
dmrs_fixed.int = countTest2(DS = ds_50_100, 
                            minCountPerIndv = 8,
                            maxGrpCV = c(1, 1),
                            Minlog2FC = 1,
                            CountPerBp = 0.01, 
                            test = 'LRT',
                            num.cores = 4L, 
                            verbose = TRUE)

dmrs_fixed.int
```

The DMR associated genes at a $distance \leq 1000$:
```{r dist_dmrs_fixed.int}
hits <- distanceToNearest(dmrs_fixed.int, GENES, ignore.strand=FALSE)

## DMR associated genes 
hits <- hits[mcols(hits)[,1] <= 1000]
GENES[subjectHits(hits)]
```


DMRs derived method = "relaxed" from dmpClusters function 
```{r dmrs_relaxed, message=FALSE, results='hide'}
.dmrs_relaxed = countTest2(DS = .ds_50_100, 
                           minCountPerIndv = 8,
                           maxGrpCV = c(1, 1),
                           Minlog2FC = 1,
                           CountPerBp = 0.01, 
                           test = 'LRT',
                           num.cores = 4L, 
                           verbose = TRUE)
.dmrs_relaxed
```

The DMR assciated genes at a $distance \leq 1000$:
```{r dist.dmrs_relaxed}
hits <- distanceToNearest(.dmrs_relaxed, GENES, ignore.strand=FALSE)

## DMR associated genes 
hits <- hits[mcols(hits)[,1] <= 1000]
GENES[subjectHits(hits)]
```


DMRs derived method = "fixed.int" from dmpClustering function 
```{r dmrs.fixed.int, message=FALSE, results='hide'}
dmrs.fixed.int = countTest2(DS = ds.50.100,
                            minCountPerIndv = 8,
                            maxGrpCV = c(1, 1),
                            Minlog2FC = 1,
                            CountPerBp = 0.01,
                            test = 'LRT',
                            num.cores = 4L,
                            verbose = TRUE)
dmrs.fixed.int
```

The DMR assciated genes at a $distance \leq 1000$:
```{r dist_dmrs.fixed.int}
hits <- distanceToNearest(dmrs.fixed.int, GENES, ignore.strand=FALSE)

## DMR associated genes 
hits <- hits[mcols(hits)[,1] <= 1000]
GENES[subjectHits(hits)]
```


DMRs derived method = "relaxed" from dmpClustering function 
```{r dmrs.relaxed, message=FALSE, results='hide'}
dmrs.relaxed = countTest2(DS = ds_50.100,
                          minCountPerIndv = 8,
                          maxGrpCV = c(1, 1), 
                          Minlog2FC = 1,
                          CountPerBp = 0.01,
                          test = 'LRT',
                          num.cores = 4L,
                          verbose = TRUE)
dmrs.relaxed
```

The DMR assciated genes at a $distance \leq 1000$:
```{r dist_dmrs.relaxed}
hits <- distanceToNearest(dmrs.relaxed, GENES, ignore.strand=FALSE)

## DMR associated genes 
hits <- hits[mcols(hits)[,1] <= 1000]
GENES[subjectHits(hits)]
```

# Naturally ocurring regions

As was mentioned in the introduction algorithmically identified DMRs do not
necessarily will be associated to the local DNA physico-chemical properties and
to local molecular evolutionary features of every DMR detected.

All the weight of the evolutionary pressure lie down on naturally ocurrying
methylation patternings inside, for example (but not limited to), exons and
introns.

For introns we have:
```{r exons_introns}
## ==== Setting up the experiment design to test for DMRs ===
nams <- names(dmps)
dmps_at_exons <- getDIMPatGenes(GR = dmps, GENES = exons,
                                ignore.strand = TRUE,
                                output = "GRanges")

## Build a RangedGlmDataSet object
ds_exons <- glmDataSet(GR = dmps_at_exons, colData = colData)

```

Differentially methylated exons
```{r dmrs_exons}
dmrs_exons = countTest2(DS = ds_exons,
                        minCountPerIndv = 8,
                        maxGrpCV = c(1, 1), 
                        Minlog2FC = 1,
                        CountPerBp = 0.01, 
                        test = 'LRT',
                        num.cores = 4L,
                        verbose = TRUE)

dmrs_exons
```

For introns we have:
```{r design_introns}
## ==== Setting up the experiment design to test for DMRs ===
nams <- names(dmps)
dmps_at_introns <- getDIMPatGenes(GR = dmps, GENES = introns,
                                 ignore.strand = TRUE,
                                 output = "GRanges")

## Build a RangedGlmDataSet object
ds_introns <- glmDataSet(GR = dmps_at_introns, colData = colData)

```

Differentially methylated introns
```{r dmrs_introns}
dmrs_introns = countTest2(DS = ds_introns,
                          minCountPerIndv = 8,
                          maxGrpCV = c(1, 1), 
                          Minlog2FC = 1,
                          CountPerBp = 0.01,
                          test = 'LRT',
                          num.cores = 4L,
                          verbose = TRUE)
dmrs_introns
```


### Differentially methylated genes DMGs

The last results for introns and exons contrasts with the result for the whole
gene-body.

```{r design_genes}
## ==== Setting up the experiment design to test for DMRs ===
nams <- names(dmps)
dmps_at_genes <- getDIMPatGenes(GR = dmps, GENES = genes,
                                ignore.strand = TRUE,
                                output = "GRanges")

## Build a RangedGlmDataSet object
ds_genes <- glmDataSet(GR = dmps_at_genes, colData = colData)

```

Differentically methylated genes
```{r dmrs_genes}
dmrs_genes = countTest2(DS = ds_genes, 
                        minCountPerIndv = 8,
                        maxGrpCV = c(1, 1),
                        Minlog2FC = 1,
                        CountPerBp = 0.01,
                        test = 'LRT',
                        num.cores = 4L, 
                        verbose = TRUE)

dmrs_genes
```

It seems to be that methylation patterning inside of gene-body forms a sort
continuous regions simultaneously covering portions of exon and introns. In
particular, this is true for genes experiencing alternative splicing, where some
regions inside introns works as exons as well.

# References.

<div id="refs"></div>


# Session info {.unnumbered}

Here is the output of `sessionInfo()` on the system on which this document was
compiled running pandoc `r rmarkdown::pandoc_version()`:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
