---
title: "Association Between Gene Expression and Cytosine DNA Methylation at gene-body"
date: "`r format(Sys.time(), '%d %B %Y')`"
fontsize: 11pt
fontfamily: "serif"
bibliography: bibliography.bib
output:
    rmarkdown::html_vignette: 
        toc: true
        toc_depth: 3
        number_sections: true
        geometry: margin=0.8in
        keep_md: yes

abstract: |
    This document is Supplementary Information for the manuscript: _"Segregation
    of an MSH1 RNAi transgene produces heritable non-genetic memory in
    association with methylome reprogramming"_ [@Yang2020]. Statistical
    analysis supporting the association between gene expression and cytosine DNA
    methylation at gene-body is provided.
vignette: >
    %\VignetteIndexEntry{Association Between Gene Expression and Cytosine DNA Methylation at gene-body}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}  

---

<style type="text/css">

.list-group-item.active, .list-group-item.active:focus, 
.list-group-item.active:hover {
    z-index: 2;
    color: #fff;
    background-color: #337ab7;
    border-color: #337ab7;
}

.tocify-subheader .tocify-item {
  font-size: 0.80em;
  padding-left: 25px;
  text-indent: 0;
}

.tocify-header .tocify-item {
  font-size: 0.85em;
  padding-left: 20px;
  text-indent: 0;
}

body{ /* Normal  */
      font-size: 18px;
      font-family: "Times New Roman", Times, serif;
      text-align: justify
  }
td {  /* Table  */
  font-size: 8px;
}

h1.title {
  font-size: 38px;
  font-family: "Times New Roman", Times, serif;
  color: DarkRed;
  .subTitle {
  font-size: 24px;
  font-family: "Times New Roman", Times, serif;
  color: DarkRed;
  }
}

h1 { /* Header 1 */
  font-size: 28px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
}
h2 { /* Header 2 */
    font-size: 22px;
    color: DarkBlue;
    font-family: "Times New Roman", Times, serif;
}
h3 { /* Header 3 */
   font-size: 18px;
   color: DarkBlue;
   font-family: "Times New Roman", Times, serif;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>

# Background
The classical statistical approach to analyze the association between two
variables is based in correlation tests. However, small or zero correlation
coefficient does not necessarily implies independence of the tested
variables/processes and, consequently, does not implies the lack of
association at all. There various steps involved in investigating the dependence
between two random variables. The state of the art include the application of
copula distributions. Herein, we provide study of the statistical dependence
structure between gene expression and methylation in memory line. To uncover
the association between gene expression and cytosine DNA methylation, first, the
probability distributions of gene expression (in terms of logarithm of fold
changes: logFC) and methylation divergences are estimated.

Next, the joint probability distribution is estimated, expressed in term of
copula distribution. All the analyses are performed in R and can be reproduced
as presented here. The association is investigated with the  application of
[copula
distributions](https://en.wikipedia.org/wiki/Copula_(probability_theory)).

# Required R packages
With the exception of _usefr_ R package, all the packages are available in Bioconductor 
and in CRAN repository. Package _usefr_ is available in GitHub at https://github.com/genomaths/usefr 
You can install usefr package from GitHub typing the following command in the R console:

```{r usefr, eval=FALSE}
 devtools::install_git("https://github.com/genomaths/usefr.git")
```

```{r packages, message=FALSE}
library( GenomicRanges )
library( ggplot2 )
library( edgeR )
library( usefr )
library( plot3D )
library( copula )
library( coin )

```


```{r, include=FALSE}
options(tinytex.verbose = TRUE)
```

# Analysis for down-regulated genes
## Data sets 

Gene expression data sets were derived using the R package edgeR. All the genes
with the minimal quality of gene expression differences are included in the
analyses. That is, to evaluate the existence of an association between gene
expression and cytosine methylation all the genes with meaningful expression and
methylation values must be included, regardless they would be differentially
methylated and/or differentially expressed or not.

Gene expression data for Arabidopsis Memory first generation (G1)

```{r meth_and_gene_exp_datasets, message=FALSE, results='hide'}
## ===== Download methylation data from Github ========
url <- paste0("github.com/genomaths/genomaths.github.io/raw/master/supp-inf/", 
              "mm_nm_wt_transgen_meth_density_and_gene-exp_at_genes_06-01-19.",
              "RData")
temp <- tempfile(fileext = ".RData")
download.file(url = url, destfile = temp)
load(temp)
file.remove(temp); rm(temp, url)

```


```{r gene_exp_data}
## Gene expresion data set
deg_mm1_wt1
```

Methylation dataset contains the densities of methylation levels from control (Col-0) and 
Memory lines samples at DMP on gene-body regions. Densities were computed using function [getGRegionsStat](https://genomaths.github.io/MethylIT_utils_HTML_Manual/getGRegionsStat-methods.html)
from the R package [MethylIT.utils](https://github.com/genomaths/MethylIT.utils).

Methylation data for a memory G1 sample
```{r M_1_105}
dmp_den_genes[, "M_1_105"]

```

## Dataset for the first generation 
The analysis will be performed with Memory line (M1) and
wild type (WT1) control, both of them, MM1 and WT1, first generation G1.
```{r first}
idx <- na.omit(match(rownames(deg_mm1_wt1), names(dmp_den_genes)))
dmp_den_m1 <- dmp_den_genes[idx, c("M_1_105", "M_1_168", "M_1_179", "M_1_27", "M_1_3")]
dmp_den_wt1 <- dmp_den_genes[idx, c( "W_1_1", "W_1_3", "W_1_4", "W_1_5")]
```

The means of methylation levels densities are computed for each group
```{r dens_means}
mcols(dmp_den_wt1) <- rowMeans(as.matrix(mcols(dmp_den_wt1)))
mcols(dmp_den_m1) <- rowMeans(as.matrix(mcols(dmp_den_m1)))
```

The absolute difference of the group density means of methylation levels between control 
(Col-0) and Memory lines at gene-body regions (G1). 
```{r dens_dif}
dmp_den_wt1_mm1 <- dmp_den_m1
mcols(dmp_den_wt1_mm1) <- as.matrix(mcols(dmp_den_m1)) - as.matrix(mcols(dmp_den_wt1))
colnames(mcols(dmp_den_wt1_mm1)) <- "dmp_den"
idx <- which(dmp_den_wt1_mm1$dmp_den > 0)
dmp_den_wt1_mm1 <- dmp_den_wt1_mm1[idx]
dmp_den_wt1_mm1
```

Final variables: |_group means density differences_| and $|logFC|$. For the sake
of brevity and clarity, we shall call the first variable: _divergence of
methylation densities_ or simply _methylation divergence_ ($MD$).
```{r}
.deg_mm1_wt1 <- deg_mm1_wt1[match(names(dmp_den_wt1_mm1), rownames(deg_mm1_wt1)), ]

MD <- dmp_den_wt1_mm1$dmp_den # Methylation (|group mean density differences|)
idx <- which(.deg_mm1_wt1$table$logFC < 0 ) # Down-regulated
logFC <- abs(.deg_mm1_wt1$table$logFC[idx]) # Gene Expression (|logFC|)
MD <- MD[idx]
DataFrame(MD,logFC)
```

There are 13962 genes to be included in the downstream analysis. 

## Distribution of the divergence of methylation level density at gene-body. MM1 vs WT1 

Searching for the best fitted distribution functions by using R functions 
[fitdistr](https://www.rdocumentation.org/packages/MASS/versions/7.3-51.4/topics/fitdistr)
from [MASS](https://www.rdocumentation.org/packages/MASS/versions/7.3-50) R package and [fitCDF](https://genomaths.github.io/usefr_manual/fitCDF.html) from
[usefr](https://github.com/genomaths/usefr) R package did not yield a suitable
results. The best fitted model was found for a mixture distribution model obtained 
with the function [fitMixDist ](https://genomaths.github.io/usefr_manual/fitMixDist.html)
from [usefr](https://github.com/genomaths/usefr) R package.

The best fitted model for $MD$, from comparison MM1 _vs_ WT1, was a mixture of Log-Normal and Weibull
distributions.
```{r meth_den_fit, warning=FALSE }
x <- MD[ MD > 0] # Weibull distribution is defineed for MD > 0.

meth_FIT <- fitMixDist(X = x, args = list(lnorm = c(meanlog = 2.3, sdlog = 0.7),
                                     weibull = c(shape = 0.5, scale = 2)),
                  dens = TRUE, npoints = 900)
summary(meth_FIT$fit)
cat("\n Mixture proportions (phi):", meth_FIT$phi)
```

The graphic for histogram and density distribution curves: 
```{r graphic_meth}
par(mar = c(4, 4, 2, 1), font.lab=2, font=2,font.axis=2,family="serif" , lwd = 1.2, cex = 1.4)
hist(x, 240, freq = FALSE, las = 1, #ylim = c(0, 0.021), cex = 1,
     panel.first={points(0, 0, pch=16, cex=1e6, col="grey95")
       grid(col="white", lty = 1)}, main = text,
     cex.main = 0.9, xlim = c(0, 5), family = "serif",
     col = "cyan1", border = "deepskyblue", xaxt ="n", yaxt = "n", ann = FALSE)
axis(1, padj = -1, las = 1, lwd = 0.8, tck = -0.02)
axis(2, hadj = 0.7, las = 1, lwd = 0.8, tck = -0.02)
mtext(side = 1, text = "Divergence of methylation level density at gene-body",
      line = 1.6, cex = 1.4, font = 3)
mtext(side = 2, text = "Density distribution", line = 2.1, cex = 1.4, font = 3)
mtext("(Log-Normal & Weibull Mixture Dists.)", cex = 1.4, font = 3)
x1 <- seq(0, 10, by = 0.01)
lines(x1, dmixtdistr(x1, phi = meth_FIT$phi, arg = meth_FIT$args), col = "blue")

```


## Distribution of the $|logFC|$. MM1 vs WT1 

The best fitted model for the $|logFC|$, from comparison MM1 _vs_ WT1, was a
mixture of Gamma and Weibull distributions. Function
[fitMixDist](https://genomaths.github.io/usefr_manual/fitMixDist.html) calls the
mixture function models set up in function
[mixtdistr](https://genomaths.github.io/usefr_manual/mixtdistr.html)

```{r logFC_fit, warning=FALSE}
x <- logFC[ logFC > 0 ]

gexp_FIT <- fitMixDist(X = x, args = list(gamma = c(shape = 2, scale = 0.41),
                                          weibull = c(shape = 2, scale = 0.4)),
                       dens = TRUE, npoints = 900)
summary(gexp_FIT$fit)
cat("\n Mixture proportions (phi):", gexp_FIT$phi)
```

The graphic for histogram and density distribution curves: 
```{r logFC_grh}
par(mar = c(3, 3, 2, 1), font.lab=2, font=2,font.axis=2,family="serif", lwd = 1.2, cex = 1.4)
hist(x, 200, freq = FALSE, las = 1, #ylim = c(0, 0.021), cex = 1,
     panel.first={points(0, 0, pch=16, cex=1e6, col="grey95")
       grid(col="white", lty = 1)}, main = text, 
     cex.main = 0.9, xlim = c(0, 4), family = "serif",
     col = "cyan1", border = "deepskyblue", xaxt ="n", yaxt = "n", ann = FALSE)
axis(1, padj = -1.5, las = 1, lwd = 0.5, tck = -0.02)
axis(2, hadj = 0.4, las = 1, lwd = 0.5, tck = -0.02)
mtext(side = 1, text = expression(italic(abs(logFC))),
      line = 1.3, cex = 1.4, font = 3)
mtext(side = 2, text = "Density distribution", line = 1.6, cex = 1.4, font = 3)
mtext("(Gamma & Weibull Mixture Dists.)", cex = 1.4, font = 3)
x1 <- seq(0, 10, by = 0.01)
lines(x1, dmixtdistr(x1, phi = gexp_FIT$phi, arg = gexp_FIT$args), col = "blue")
```


## Joint distribution of $MD$ and $|logFC|$ expressed in terms of Copula distribution.

There various steps involved in investigating the dependence between two random
variables. The state of the art in this subject include the application of
copula distributions.

Sklar's theorem ([@Sklar1959; @Sklar1973]) named after Abe Sklar, provides the 
theoretical foundation for 
the application of copulas. Sklar's theorem states that every multivariate 
cumulative distribution function (CDF):

$$H(x_1, ..., x_d)=Pr(X_1 \leq x_1, ..., X_d \leq x_d)$$ 

of any $d-tuple$ of continuous random variables $(X_1, ..., X_d)$ can be 
expressed in terms of its marginals CDFs: 
$F_i(x_i)=Pr(X_i \leq x_i)$ a copula $C$:

$$H(x_1, ..., x_d)=C(F_1(x_1), ..., F_d(x_d))$$
While Sklar showed that $C$ and $F_i$ are uniquely determined when H is known, a
valid model for $(X_1, ..., X_d)$ arises from the last equation whenever the
"ingredients" are chosen from given parametric families of distribution
([@Christian2007]): $$F \in (F_\delta), C \in (C_\alpha)$$

Current literature on copula theory provides abundant information on models and
copula families. Here, for the particular case of bi-dimensional copula defined
for $MD$ and $|logFC|$, we have chosen Farlie–Gumbel–Morgenstern family of
copulas, which are defined for, each $\theta \in [-1, 1]$ by: $$C_\alpha(u,
v)=uv+\alpha uv(1-u)(1-v)$$ $u,v \in [0,1]$. The correlation coefficient $\rho$
is linked to $\alpha$ by the relationship: $\rho=\frac{\alpha}{3}$, which ranges
from $-\frac{1}{3}$ to $\frac{1}{3}$, where $u=F(x)$, $v=G(y)$, with marginal
CDFs: $F(x)$ and $G(y)$.

### Fitting Farlie–Gumbel–Morgenstern copulas with $MD$ and $|logFC|$ marginal CDFs

The first step requires for provide the parameters values, previously estimated, of the marginal distributions, which
were given throughout the function [mixtdistr](https://genomaths.github.io/usefr_manual/mixtdistr.html). The
the specific parameter values are retrieved from the previous nonlinear fit obtained for the marginal CDFs:
```{r cop_fit_par}
margins = c("mixtdistr", "mixtdistr")
parMargins = list(
  list(phi = meth_FIT$phi, arg = meth_FIT$args),
  list(phi = gexp_FIT$phi, arg = gexp_FIT$args))
```

The parameter estimation and goodness-of-fit (GOF) test take advantages of the tools
available in the R package [copula](https://www.rdocumentation.org/packages/copula/versions/0.999-19.1).
Both, parameter estimation and GOF, can be accomplished with function 
[bicopulaGOF](https://genomaths.github.io/usefr_manual/bicopulaGOF.html) from the
R package [usefr](https://github.com/genomaths/usefr).

```{r cop_fit, message = FALSE, warning=FALSE, comment=FALSE}
fgm_cop <- bicopulaGOF(x = MD, y = logFC, copula = "fgmCopula", sample.size = 13000, 
                       margins = margins, paramMargins = parMargins, nboots = 10, 
                       approach = "chisq", breaks = 30, num.cores = 6L, seed = 123,
                       verbose = FALSE)
```

The null hypothesis that the data follows FGM copula distribution is not rejected.
```{r res}
fgm_cop$gof
```


All the information needed about the fitted FGM copula is found in the object
_fgm_cop$copula_, which is a [mvdc](https://www.rdocumentation.org/packages/copula/versions/0.999-19.1/topics/mvdc-class)
object from [copula](https://www.rdocumentation.org/packages/copula/versions/0.999-19.1)
package with several slots. In particular, the specific information about FGM copula is found in:

```{r fgm_fitted}
fgm_cop$copula@copula
```

### P-P plot. Graphical Diagnostics of GOF
The P-P plot is build with function [ppCplot](https://genomaths.github.io/usefr_manual/ppCplot.html)
from the R package [usefr](https://github.com/genomaths/usefr). If the GOF is good enough, then it 
is expected that points in the graphics are closed to the diagonal (red) straight line. Details on
how the computation is performed are provided in the help for function. [ppCplot](https://genomaths.github.io/usefr_manual/ppCplot.html).

```{r ppPlot}
g <- ppCplot(X = MD, Y = logFC, copula = fgm_cop$copula, npoints = 500, col = "blue", cex = 0.5)
```


Information on the data and copula used to yield the graphic are stored in the
output variable _"g"_. For example, the empirical and theoretical probabilities used
to built the graphic can be retrieved from the data frame: _"g$data"_:

```{r g_data}
DataFrame(g$data)
```

### Three-dimensional copula density plot

The domain region $(X, Y)$ is defined and function [dMvdc](https://www.rdocumentation.org/packages/copula/versions/0.999-19.1/topics/Mvdc)
is used to compute the density $z$ in the grid $[X,Y] \times [X,Y]$ using the 
previous FMG fitted copula.
```{r domain}
X = seq(0, 1.5, length = 100)
Y = seq(0, 1.5, length = 100)
sp_rho <- round(rho(fgm_cop$copula@copula),3) # Spearman's rho: -0.171


z <- outer(X, Y, function(x,y) dMvdc(cbind(x,y),fgm_cop$copula)) * 4
phi = 20; theta = 45

zf <- function(x,y) dMvdc(cbind(x,y),fgm_cop$copula) * 3.2

cat("Spearman's rho = ", sp_rho)

```

An auxiliary function to draw the panels 
```{r panels}
panelfirst <- function(pmat) {
  par(mfrow = c(1, 1),mar=c(4,7,0, 1), font=2,lwd = 0.01, cex =1, family = "serif")
  YZ <- trans3D(x = rep(0, ncol(z)), Y, z[18,], pmat = pmat)

  scatter2D(YZ$x, YZ$y, colvar = z[18,], lwd = 1, 
            type = "l", add = TRUE, colkey = FALSE)

  XZ <- trans3D(x = X, y = rep(max(Y), nrow(z)),
                z = z[, 12], pmat = pmat)

  scatter2D(XZ$x, XZ$y, colvar = z[, 12], lwd = 1, 
            type = "l", add = TRUE, colkey = FALSE)
  
}
```

The graphical settings depend on the final printing destiny and on the dataset.
```{r 3d_cop, message=FALSE, message=FALSE}
par(mfrow = c(1, 1),mar=c(4,7,0, 1), font=2,lwd = 0.1, cex =1, family = "serif")
persp3D(X, Y, z, lwd = 0.01, image = TRUE,
        xlab = "", ylab = "", zlab = "", family = "serif",
        box = T, panel.first = panelfirst, bty = "u",
        col.panel = "steelblue", col.grid = "black", lwd.grid = 0.5,
        contour = list(col = "black", side = "zmin", lwd = 0.4),
        ticktype = "detailed", clab = "", expand = 0.7, d = 2,
        phi = phi, theta = theta, resfac = 2.5, cex.lab = 1, lwd = 0.005,
        xlim = c(0, 1.5), zlim = c(-5, 6), font=2, cex.axis = 1.2,
        colkey = list(length = 0.7, width = 0.6, col.ticks = "black", lwd = 0.01,
                      font=2, shift = 0, dist = 0, lwd.ticks = 0.01,  cex.clab = 1.2,
                      cex.axis = 1, line = -0.8))
text(0.38,-0.48, srt = 34, labels = expression(paste("|", italic(log[2]), italic(FC), "|")),
     xpd = TRUE,font = 2, cex = 1.2)
text(-0.3,-0.44, srt = -28, labels = expression(paste(italic("MD"))),
     xpd = TRUE, font=2, cex = 1.2)
text(-0.3,-0.6, srt = 0, labels = expression(paste("Spearman' s ", italic(rho), ": ", -0.17)),
     xpd = TRUE, font = 1, cex = 1)
```


The marginal distribtuions are drawn in the $XZ$ and $YZ$ vertical planes.
Consistent with the Spearman' s $\rho=-0.17$, for a fixed value of  $|log_2FC|$, 
(a line parallel to $MD$ axis) the joint probability density distribution 
function $f(X \leq MD, Y \leq |log_2FC|)$ decreases with the increment of $MD$.
That is, the probability to observe DEGs with values, say between 0.1 and 0.5,  
$0.1 \leq |log_2FC| \leq 0.5$, is greater for $0.1 \eq MD \leq 0.5$ than for 
$0.5 < MD \leq 1$.

### Two-dimensional copula density plot, Contour plot
The 2D-dimensional plot permits to see in more detail the projection of the 
density surface on the $XY$ plane and contour lines.

```{r contour, echo=FALSE}
X = seq(-0, 2, length = 300)
Y = seq(-0, 2, length = 300)
z <- outer(X, Y, function(x,y) dMvdc(cbind(x,y),fgm_cop$copula)) * 3.2

par(family = "serif", font  = 2, bty="n", mar = c(4, 4, 2, 1.1))
image2D(z = z, x = X, y = Y, colvar = z,  col = NULL, NAcol = "white", breaks = NULL,
        border = NA, facets = TRUE, xlab = "", ylab = "", ylim = c(-0.1, 1.3),
        xlim = c(0, 2), xaxt ="n", yaxt = "n", ann = FALSE,
        resfac = 1, clab = NULL, ticktype = "detailed",
        colkey = list(length = 0.9, width = 0.7, col.ticks = "black", lwd = 0.01,
                      font=2, shift = 0.02, dist = 0, lwd.ticks = 0.1,  cex.clab = 1.2,
                      cex.axis = 1, line = -0.8, tcl = -1),
        lighting = FALSE, shade = NA, ltheta = -135, lphi = 0,
        theta = 0, rasterImage = FALSE,
        add = FALSE, plot = TRUE)
contour(x = X, y = Y, z = z, add = TRUE, ylab = "logFC", nlevels = 11, bty = "n",
        xlim = c(-0.05, 3), ylim = c(-0.05, 1.5), xaxt ="n", yaxt = "n", ann = FALSE,
        lwd = 0.4, labcex = 0.9, vfont = c("serif", "bold"))
axis(1, padj = -1.2, las = 1, lwd = 0.5, tck = -0.02, line = -0.6)
axis(2, hadj = 0.4, las = 1, lwd = 0.5, tck = -0.02, line = 0.6)
mtext(side = 1, text = expression(paste(italic("MD"))),
      line = 1.3, cex = 1.4, font = 3)
mtext(side = 2, text = expression(italic(abs(logFC))), line = 2.5, cex = 1.4, font = 3)
abline(v = 0.1, col = "black", lty = "dashed")
abline(v = 0.5, col = "black", lty = "dashed")
abline(v = 1, col = "black", lty = "dashed")
abline(h = 0.1, col = "black", lty = "dashed")
abline(h = 0.5, col = "black", lty = "dashed")
abline(h = 1, col = "black", lty = "dashed")
```

It is not difficult to see that for a fixed value of $MD$ (a line parallel to $|log_2FC|$ axis) 
the joint probability density distribution function $f(X \leq MD, Y \leq |log_2FC|)$ 
(color gradient from red to blue) decreases with the increment of $|log_2FC|$. 
That is, the probability to observe $MD$ values, say between 0.1 and 0.5
(see the squared region below) $0.1 \leq MD \leq 0.5$, is greater for 
$0.1 \leq |log_2FC| \leq 0.5$ than for $0.5 < |log_2FC| \leq 1$.

In summary, gene expression and methylation processes (2kb upstream gene TSS
regions) are not independent in memory line first generation. There exist an
stochastic deterministic dependence between $|log_2FC|$ and $MD$ that can be
expressed in terms joint probability distribution, which leads to an inverse
relationship between the conditioned expected values $|log_2FC|$ given $MD$, and
vice versa. This association is, however, weaker than the association found
for gene expression and methylation at gene body.

## Asymptotic linear-by-linear association test
The above 3D and 2D density plots and results suggest that most of the genes 
have low $|logFC|$ and low $MD$, and the number of genes decrease for middle
levels of $|logFC|$ and $MD$. This observation leads to the application of a
linear-by-linear association test.

The linear-by-linear test (LLT) is used to test a linear association among
variables in a contingency table with ordered categories (@Agresti2007). The
null hypothesis for the linear-by-linear test is that there is no association
among the variables in the table.

It is obvious from the 2D density plots that a relationship for $MD$ < 0.08 
is not meaningful. In particular, both variables, $|logFC|$ and $MD$, are at the 
level of technical noise. The percentage of the genes at noise level are:
```{r noise}
round(100 * sum(MD < 0.08)/length(MD), 2)
```

Hence,  the genes considered in the downstream analysis are:
```{r filtering}
idx <- which(MD > 0.08)
md  <- MD[ idx]
logfc  <- logFC[ idx ]
dt <- data.frame(md = md , logfc = logfc)
nrow(dt)
```

### Discretization of $MD$ and $|logFC|$

To apply the LLT the values of $|logFC|$ and $MD$ must be discretized into 
ordinal values. A K-means discretization approach is applied with following 
auxiliary function

```{r kmeanf}
# ------------------- Auxiliary function for discretization ------------------ #
Kmean.discretization <- function(x, categories, algorithm = "Hartigan-Wong",
                                 iter.max = 10^6, column = 1L) {
  centers <- kmeans(x, centers = categories, algorithm = algorithm,
                    iter.max = iter.max)$centers
  # order the centers
  centers <- sort(centers)
  cl <- stats::kmeans(stats::na.omit(x), centers = centers,
                      algorithm = algorithm, iter.max = iter.max )
  return( data.frame(x, cluster = cl$cluster) )
}
```


### Discretization for $MD$
```{r disc_md}
cl = Kmean.discretization(md, categories = 3, algorithm = "MacQueen" )
cl_summary <- sapply(1:3, function(k) summary(cl$x[cl$cluster == k]))
colnames(cl_summary) <- c("low", "middle", "high")
cl_summary
```

Cutpoint are now applied:
```{r md_cut}
cutpoint1 <- min(cl_summary[,2])
cutpoint2 <- min(cl_summary[,3])

dt$md_level <- rep("High")
dt$md_level[ dt$md < cutpoint2 ] <- "Middle"
dt$md_level[ dt$md < cutpoint1   ] <- "Low"
dt$md_level <- factor(dt$md_level, levels = c("Low", "Middle", "High"))
```

### Discretization for $|logFC|$

```{r disc_logfc}
cl = Kmean.discretization(logfc, categories = 3, algorithm = "MacQueen" )
cl_summary <- sapply(1:3, function(k) summary(cl$x[cl$cluster == k]))
colnames(cl_summary) <- c("eLow", "eMiddle", "eHigh")
cl_summary
```

```{r}
cutpoint1 <- min(cl_summary[,2])
cutpoint2 <- min(cl_summary[,3])

dt$logfc_level <- rep("eHigh")
dt$logfc_level[ dt$logfc < cutpoint2 ] <- "eMiddle"
dt$logfc_level[ dt$logfc < cutpoint1   ] <- "eLow"

dt$logfc_level <- factor(dt$logfc_level, levels = c("eLow", "eMiddle", "eHigh"))

```


### The contingency table and its graphic visualization

In the current case, the contingency table suggests an obvious relationship between
$|logFC|$  and $MD$.
```{r contbl}
m <- as.matrix( base::table( dt$logfc_level, dt$md_level) )
m
```

The [spineplot](https://www.rdocumentation.org/packages/graphics/versions/3.6.0/topics/spineplot)
of this contingency table (table(x = logfc_level, y = md_level)) help to visualize the
existent relationship between $|logFC|$ and $MD$.

```{r spineplot, fig.height=6, fig.width=8}
par( mar = c(5, 7, 4, 4), family = "Serif", font = 2, cex = 0.8, las = 1, lwd = 0.2 )
tb <- spineplot(m[, c(3,2,1)], col = c( "blue", "dodgerblue","deepskyblue" ),
                xlab = "", ylab = "")
mtext(side = 1, text = expression(italic(abs(logFC))), 
      line = 2.5, cex = 1, font = 3)
mtext(side = 2, text = expression(paste(italic("MD"))),
      line = 4, cex = 1, font = 3)
abline(a = 0.0, b = 0.72, col = "white", lty = "dashed", lwd = 1)
```

A dashed line was added to emphasize the direction of linear trend. From _"Low"_ to 
_"high"_ there seems to be a linear relationship between pairwise proportion of
$|logFC|$ and $MD$.

Conceptually, the _spineplot_ is a plot of the conditional probability $P(y|x)$
against $P(x)$. Since the joint probability distribution of these variable is
described by a FGM copula, a linear relationship between $P(y|x)$ against $P(x)$
must be expected, which is a property of FGM copula [@Balakrishna2009].

### Testing association between $|logFC|$ and $MD$
Here, we use the LLT test implemented in function  [lbl_test](https://www.rdocumentation.org/packages/coin/versions/1.3-0/topics/ContingencyTests)
from the R package _coin_. Two test variants are available: 1) permutation test 
and asymptotic test. 

The asymptotic LLT test:
```{r allt}
lbl_test( m )
```


The permutation LLT test:
```{r pllt}
lbl_test( m, distribution = "approximate", B = 10000, alternative = "less" )
```
The null hypothesis of no association among the variables in the table is
rejected. In the current case the p-value < 1e-04 and, hence, we do not 
have enough evidence to support the null hypothesis.

## Generalized linear model (GLM)

The variable $MD$ was split into three levels, "0": "Low" ($0 < MD \leq 0.5$),
"1": "Middle" ($0.5 < MD \leq 1$), and "2": "High" ($1 < MD $)


```{r glm_dt_3}
meth_cut <- 1

xy <- cbind(MD, logFC)
idx <- which(logFC > 0.5)
xy <- data.frame(xy[idx, ])
xy$md <- 0
xy$md[xy$MD > meth_cut] <- 1
```

The glm support the decrease of $-logFC$ with th increment of $MD$. The model
coefficients are highly statistically significant.
```{r glm}
gm <- glm(formula = logFC ~ md, data=xy, family = Gamma(link = "identity"))
summary( gm )
```

The ANOVA test confirm the negative trend given by the coefficient of MD 
(highly statistically significant).
```{r anova_gamma}
anova(gm, test = "LRT")

```


### Boxplot
The boxplot suggest a clear trend of $-logFC$ decreasing with increment of MD
levels. Notice that the _notchs_ from the boxes do not overlap, which implies
that the differences between medians from three groups ("Low", "Middle", and
"High") are statistically significant. So, the GLM results can be visualized.

```{r boxplot_2, fig.width=9, fig.height=6}

par(bty = "n", family = "serif", font = 2, cex = 1, lwd = 0.1, mar = c(4, 5, 2, 1.1))
boxplot(logFC ~ md, data = xy, ylim = c(0.4, 2.5), xaxt ="n",  yaxt = "n", ann = FALSE)
rect(0.5, 0.4, 3.5, 5, col = "grey94", lty = 0, lwd = 1)
grid(0, 5, col = "white", lwd = 2, lty = 1)
boxplot(logFC ~ md, data = xy, ylim = c(0.4, 2.5), notch = TRUE, xaxt ="n",
        yaxt = "n", ann = FALSE, pch = 20, medcol = "darkblue", whisklwd = 1, staplelwd = 1,
        col = c("deepskyblue", "dodgerblue"), outcol = "red", add  = TRUE)
axis(1, padj = 0.1, at = 1:2, lwd = 0.5, tck = -0.02, line = -0.6, cex.axis = 1.3, 
     labels = c("Low", "High"))
axis(2, hadj = 1, las = 1, lwd = 0.5, tck = -0.02, line = -1, cex.axis = 1.4)
mtext(side = 1, text = expression(paste(italic("MD"))),
      line = 2.2, cex = 1.4, font = 3)
mtext(side = 2, text = expression(italic(-logFC)), line = 2.1, cex = 1.4, font = 3)
text(x = 1.5, y = 2.2, "***", pos = 3, cex = 1.2 )

```

# Analysis for up-regulated genes
## Data set (II)

```{r}
.deg_mm1_wt1 <- deg_mm1_wt1[match(names(dmp_den_wt1_mm1), rownames(deg_mm1_wt1)), ]

MD <- dmp_den_wt1_mm1$dmp_den # Methylation (|group mean density differences|)
idx <- which(.deg_mm1_wt1$table$logFC > 0 ) # Up-regulated
logFC <- .deg_mm1_wt1$table$logFC[idx] # Gene Expression (|logFC|)
MD <- MD[idx]
DataFrame(MD,logFC)
```

There are 6060 genes with $MD > 0$ and $logFC > 0$ to be included in the 
downstream analysis. 

## Distribution of the divergence of methylation level density at gene-body. MM1 vs WT1 (II)

The best fitted model for $MD$, from comparison MM1 _vs_ WT1, was a mixture of
Log-Normal and Weibull distributions.
```{r meth_den_fit_2, warning=FALSE }

meth_FIT <- fitMixDist(X = MD, args = list(lnorm = c(meanlog = 2.3, sdlog = 0.7),
                                     weibull = c(shape = 0.5, scale = 2)),
                  dens = TRUE, npoints = 900)
summary(meth_FIT$fit)
cat("\n Mixture proportions (phi):", meth_FIT$phi)
```

The graphic for histogram and density distribution curves: 
```{r graphic_meth_2, fig.height=6, fig.width=8}
par(mar = c(4, 4, 2, 1), font.lab=2, font=2,font.axis=2,family="serif" , lwd = 1.2, cex = 1.4)
hist(MD, 90, freq = FALSE, las = 1, #ylim = c(0, 0.021), cex = 1,
     panel.first={points(0, 0, pch=16, cex=1e6, col="grey95")
       grid(col="white", lty = 1)}, main = text,
     cex.main = 0.9, xlim = c(0, 5), family = "serif",
     col = "cyan1", border = "deepskyblue", xaxt ="n", yaxt = "n", ann = FALSE)
axis(1, padj = -1, las = 1, lwd = 0.8, tck = -0.02)
axis(2, hadj = 0.7, las = 1, lwd = 0.8, tck = -0.02)
mtext(side = 1, text = "Divergence of methylation level density at gene-body",
      line = 1.6, cex = 1.4, font = 3)
mtext(side = 2, text = "Density distribution", line = 2.1, cex = 1.4, font = 3)
mtext("(Log-Normal & Weibull Mixture Dists.)", cex = 1.4, font = 3)
x1 <- seq(0, 10, by = 0.01)
lines(x1, dmixtdistr(x1, phi = meth_FIT$phi, arg = meth_FIT$args), col = "blue")

```


## Distribution of the $|logFC|$. MM1 vs WT1 (II)

The best fitted model for the $|logFC|$, from comparison MM1 _vs_ WT1, was a
mixture of Gamma and Weibull distributions. Function
[fitMixDist](https://genomaths.github.io/usefr_manual/fitMixDist.html) calls the
mixture function models set up in function
[mixtdistr](https://genomaths.github.io/usefr_manual/mixtdistr.html)

```{r logFC_fit_2, warning=FALSE}
gexp_FIT <- fitMixDist(X = logFC, args = list(gamma = c(shape = 2, scale = 0.41),
                                          weibull = c(shape = 2, scale = 0.4)),
                       dens = TRUE, npoints = 900)
summary(gexp_FIT$fit)
cat("\n Mixture proportions (phi):", gexp_FIT$phi)
```

The graphic for histogram and density distribution curves: 
```{r logFC_grh_2, fig.height=6, fig.width=8}
par(mar = c(3, 3, 2, 1), font.lab=2, font=2,font.axis=2,family="serif", lwd = 1.2, cex = 1.4)
hist(logFC, 120, freq = FALSE, las = 1, #ylim = c(0, 0.021), cex = 1,
     panel.first={points(0, 0, pch=16, cex=1e6, col="grey95")
       grid(col="white", lty = 1)}, main = text, 
     cex.main = 0.9, xlim = c(0, 4), family = "serif",
     col = "cyan1", border = "deepskyblue", xaxt ="n", yaxt = "n", ann = FALSE)
axis(1, padj = -1.5, las = 1, lwd = 0.5, tck = -0.02)
axis(2, hadj = 0.4, las = 1, lwd = 0.5, tck = -0.02)
mtext(side = 1, text = expression(italic(abs(logFC))),
      line = 1.3, cex = 1.4, font = 3)
mtext(side = 2, text = "Density distribution", line = 1.6, cex = 1.4, font = 3)
mtext("(Gamma & Weibull Mixture Dists.)", cex = 1.4, font = 3)
x1 <- seq(0, 10, by = 0.01)
lines(x1, dmixtdistr(x1, phi = gexp_FIT$phi, arg = gexp_FIT$args), col = "blue")
```


## Joint distribution of $MD$ and $|logFC|$ expressed in terms of Copula distribution (II)
### Fitting Farlie–Gumbel–Morgenstern copulas with $MD$ and $|logFC|$ marginal CDFs (II)

The first step requires for provide the parameters values, previously estimated,
of the marginal distributions, which were given throughout the function
[mixtdistr](https://genomaths.github.io/usefr_manual/mixtdistr.html). The
specific parameter values are retrieved from the previous nonlinear fit obtained
for the marginal CDFs:
```{r cop_fit_par_2}
margins = c("mixtdistr", "mixtdistr")
parMargins = list(
  list(phi = meth_FIT$phi, arg = meth_FIT$args),
  list(phi = gexp_FIT$phi, arg = gexp_FIT$args))
```

The parameter estimation and goodness-of-fit (GOF) test take advantage of the
tools available in the R package
[copula](https://www.rdocumentation.org/packages/copula/versions/0.999-19.1).
Both, parameter estimation and GOF, can be accomplished with function
[bicopulaGOF](https://genomaths.github.io/usefr_manual/bicopulaGOF.html) from
the R package [usefr](https://github.com/genomaths/usefr).

```{r cop_fit_2, message = FALSE, warning=FALSE, comment=FALSE}
fgm_cop <- bicopulaGOF(x = MD, y = logFC, copula = "fgmCopula", sample.size = 5999, 
                       margins = margins, paramMargins = parMargins, nboots = 10, 
                       approach = "chisq", breaks = 30, num.cores = 6L, seed = 123,
                       verbose = FALSE)
```

The null hypothesis that the data follows FGM copula distribution is not rejected.
```{r res_2}
fgm_cop$gof
```


All the information needed about the fitted FGM copula is found in the object
_fgm_cop$copula_, which is a
[mvdc](https://www.rdocumentation.org/packages/copula/versions/0.999-19.1/topics/mvdc-class)
object from
[copula](https://www.rdocumentation.org/packages/copula/versions/0.999-19.1)
package with several slots. In particular, the specific information about FGM
copula is found in:

```{r fgm_fitted_2}
fgm_cop$copula@copula
```

### P-P plot. Graphical Diagnostics of GOF (II)
The P-P plot is build with function
[ppCplot](https://genomaths.github.io/usefr_manual/ppCplot.html) from the R
package [usefr](https://github.com/genomaths/usefr). If the GOF is good enough,
then it is expected that points in the graphics are closed to the diagonal (red)
straight line. Details on how the computation is performed are provided in the
help for function.
[ppCplot](https://genomaths.github.io/usefr_manual/ppCplot.html).

```{r ppPlot_2, fig.height=6, fig.width=8}
g <- ppCplot(X = MD, Y = logFC, copula = fgm_cop$copula, npoints = 500, col = "blue", cex = 0.5)
```


Information on the data and copula used to yield the graphic are stored in the
output variable _"g"_. For example, the empirical and theoretical probabilities used
to built the graphic can be retrieved from the data frame: _"g$data"_:

```{r g_data_2}
DataFrame(g$data)
```

### Three-dimentional copula density plot (II)

The domain region $(X, Y)$ is defined and function [dMvdc](https://www.rdocumentation.org/packages/copula/versions/0.999-19.1/topics/Mvdc)
is used to compute the density $z$ in the grid $[X,Y] \times [X,Y]$ using the 
previous FMG fitted copula.
```{r domain_2}
X = seq(0, 1.5, length = 100)
Y = seq(0, 1.8, length = 100)
sp_rho <- round(rho(fgm_cop$copula@copula),3) # Spearman' s rho: -0.171

z <- outer(X, Y, function(x,y) dMvdc(cbind(x,y),fgm_cop$copula)) * 4
phi = 15; theta = 45
sp_rho
```

An auxiliary function to draw the panels 
```{r panels_2}
panelfirst <- function(pmat) {
  par(mfrow = c(1, 1),mar=c(4,7,0, 1), font=2,lwd = 0.01, cex =1, family = "serif")
  YZ <- trans3D(x = rep(0, ncol(z)), Y, z[18,], pmat = pmat)

  scatter2D(YZ$x, YZ$y, colvar = z[18,], lwd = 1, 
            type = "l", add = TRUE, colkey = FALSE)

  XZ <- trans3D(x = X, y = rep(max(Y), nrow(z)),
                z = z[, 12], pmat = pmat)

  scatter2D(XZ$x, XZ$y, colvar = z[, 12], lwd = 1, 
            type = "l", add = TRUE, colkey = FALSE)
  
}
```

The graphical settings depend on the final printing destiny and on the dataset.
```{r 3d_cop_2, message=FALSE, message=FALSE, fig.height=8, fig.width=8}
par(mfrow = c(1, 1),mar=c(4,7,0, 1), font=2,lwd = 0.1, cex =1, family = "serif")
persp3D(X, Y, z, lwd = 0.01, image = TRUE,
        xlab = "", ylab = "", zlab = "", family = "serif",
        box = T, panel.first = panelfirst, bty = "u",
        col.panel = "steelblue", col.grid = "black", lwd.grid = 0.5,
        contour = list(col = "black", side = "zmin", lwd = 0.4),
        ticktype = "detailed", clab = "", expand = 0.7, d = 2,
        phi = phi, theta = theta, resfac = 2.5, cex.lab = 1, lwd = 0.005,
        xlim = c(0, 1.5), zlim = c(-5, 9), font=2, cex.axis = 1.2,
        colkey = list(length = 0.5, width = 0.6, col.ticks = "black", lwd = 0.01,
                      font=2, shift = 0, dist = 0, lwd.ticks = 0.01,  cex.clab = 1.2,
                      cex.axis = 1, line = -0.8))
text(0.38,-0.4, srt = 38, labels = expression(paste("|", italic(log[2]), italic(FC), "|")),
     xpd = TRUE,font = 2, cex = 1.2)
text(-0.41,-0.39, srt = -20, labels = expression(paste(italic("MD"))),
     xpd = TRUE, font=2, cex = 1.2)
text(-0.36,-0.5, srt = 0, labels = expression(paste("Spearman' s ", italic(rho), ": ", -0.164)),
     xpd = TRUE, font = 1, cex = 1)
```


The marginal distributions are drawn in the $XZ$ and $YZ$ vertical planes.
Consistent with the Spearman' s $\rho=-0.164$, for a fixed value of  $|log_2FC|$, 
(a line parallel to $MD$ axis) the joint probability density distribution 
function $f(X \leq MD, Y \leq log_2FC)$ decreasses with the increment of $MD$.
That is, the probability to observe DEGs with values, say between 0.1 and 0.5,  
$0.1 \leq log_2FC \leq 0.5$, is greater for $0.1 \leq MD \leq 0.5$ than for 
$0.5 < MD \leq 1$.

### Two-dimensional copula density plot, Contour plot (II)
The 2D-dimensional plot permits to see in more detail the projection of the 
density surface on the $XY$ plane and contour lines.

It is not difficult to see that for a fixed value of $MD$ (a line parallel to
$|log_2FC|$ axis) the joint probability density distribution function $f(X \leq
MD, Y \leq |log_2FC|)$ (color gradient from red to blue) decreasses with the
increment of $|log_2FC|$. That is, the probability to observe $MD$ values, say
between 0.1 and 0.5 (see the squared region below) $0.1 \leq MD \leq 0.5$, is
greater for $0.1 \leq |log_2FC| \leq 0.5$ than for $0.5 < |log_2FC| \leq 1$.

```{r contour2_2, echo=FALSE, fig.height=6, fig.width=8}
X = seq(-0, 2, length = 300)
Y = seq(-0, 0.8, length = 300)
z <- outer(X, Y, function(x,y) dMvdc(cbind(x,y),fgm_cop$copula)) * 3.2

par(family = "serif", font  = 2, bty="n", mar = c(4, 4, 2, 1))
image2D(z = z, x = X, y = Y, colvar = z,  col = NULL, NAcol = "white", breaks = NULL,
        border = NA, facets = TRUE, xlab = "", ylab = "", ylim = c(0, 0.8),
        xlim = c(0, 2), xaxt ="n", yaxt = "n", ann = FALSE,
        resfac = 1, clab = NULL, ticktype = "detailed",
        colkey = list(length = 0.9, width = 0.7, col.ticks = "black", lwd = 0.01,
                      font=2, shift = 0.02, dist = 0, lwd.ticks = 0.1,  cex.clab = 1.2,
                      cex.axis = 1, line = -0.8, tcl = -1),
        lighting = FALSE, shade = NA, ltheta = -135, lphi = 0,
        theta = 0, rasterImage = FALSE,
        add = FALSE, plot = TRUE)
contour(x = X, y = Y, z = z, add = TRUE, ylab = "logFC", nlevels = 10, bty = "n",
        xlim = c(-0.05, 3), ylim = c(-0.05, 0.8), xaxt ="n", yaxt = "n", ann = FALSE,
        lwd = 0.4, labcex = 0.9, vfont = c("serif", "bold"), method = "simple")
axis(1, padj = -1.2, las = 1, lwd = 0.5, tck = -0.02, line = 0.6)
axis(2, hadj = 0.4, las = 1, lwd = 0.5, tck = -0.02, line = 0.6)
mtext(side = 1, text = expression(paste(italic("MD"))),
      line = 2, cex = 1.4, font = 3)
mtext(side = 2, text = expression(italic(logFC)), line = 2, cex = 1.4, font = 3)
abline(v = 0.1, col = "black", lty = "dashed", lwd = 0.9)
abline(v = 0.5, col = "black", lty = "dashed", lwd = 0.9)
abline(v = 1, col = "black", lty = "dashed", lwd = 0.9)
abline(h = 0.1, col = "black", lty = "dashed", lwd = 0.9)
abline(h = 0.5, col = "black", lty = "dashed", lwd = 0.9)
abline(h = 1, col = "black", lty = "dashed", lwd = 0.9)
```

In summary, gene expression and methylation processes are not independent in
memory line first generation. There exist an stochastic deterministic dependence
between $|log_2FC|$ and $MD$ that can be expressed in terms joint probability
distribution, which leads to an inverse relationship between the conditioned
expected values $|log_2FC|$ given $MD$, and vice versa.


## Asymptotic linear-by-linear association test (II)

As in the previous case (above), tt is obvious from the 2D density plots that
for $MD$ < 0.08 the any relationship is not meaningful. In particular, both
variables, $logFC$ and $MD$, are at the level of technical noise. The percentage
of the genes at noise level are:
```{r noise_2}
round(100 * sum(MD < 0.08)/length(MD), 2)
```

The genes for the downstream analysis are:
```{r filtering_2}
idx <- which(MD > 0.08)
md  <- MD[ idx]
logfc  <- logFC[ idx ]
dt <- data.frame(md = md , logfc = logfc)
nrow(dt)
```

### Discretization of $MD$ and $logFC$ (II)

```{r kmeanf_2}
# ------------------- Auxiliary function for discretization ------------------ #
Kmean.discretization <- function(x, categories, algorithm = "Hartigan-Wong",
                                 iter.max = 10^6, column = 1L) {
  centers <- kmeans(x, centers = categories, algorithm = algorithm,
                    iter.max = iter.max)$centers
  # order the centers
  centers <- sort(centers)
  cl <- stats::kmeans(stats::na.omit(x), centers = centers,
                      algorithm = algorithm, iter.max = iter.max )
  return( data.frame(x, cluster = cl$cluster) )
}
```


### Discretization for $MD$ (II)
```{r disc_md_2}
cl = Kmean.discretization(md, categories = 3, algorithm = "MacQueen" )
cl_summary <- sapply(1:3, function(k) summary(cl$x[cl$cluster == k]))
colnames(cl_summary) <- c("low", "middle", "high")
cl_summary
```

Cutpoint are now applied:
```{r md_cut_2}
cutpoint1 <- min(cl_summary[,2])
cutpoint2 <- min(cl_summary[,3])

dt$md_level <- rep("High")
dt$md_level[ dt$md < cutpoint2 ] <- "Middle"
dt$md_level[ dt$md < cutpoint1   ] <- "Low"
dt$md_level <- factor(dt$md_level, levels = c("Low", "Middle", "High"))
```

### Discretization for $logFC$ (II)

```{r disc_logfc_2}
cl = Kmean.discretization(logfc, categories = 3, algorithm = "MacQueen" )
cl_summary <- sapply(1:3, function(k) summary(cl$x[cl$cluster == k]))
colnames(cl_summary) <- c("eLow", "eMiddle", "eHigh")
cl_summary
```

```{r}
cutpoint1 <- min(cl_summary[,2])
cutpoint2 <- min(cl_summary[,3])

dt$logfc_level <- rep("eHigh")
dt$logfc_level[ dt$logfc < cutpoint2 ] <- "eMiddle"
dt$logfc_level[ dt$logfc < cutpoint1   ] <- "eLow"

dt$logfc_level <- factor(dt$logfc_level, levels = c("eLow", "eMiddle", "eHigh"))

```


### The contingency table and its graphic visualization (II)

In the current case, the contingency table suggests an obvious relationship between
$|logFC|$  and $MD$.
```{r contbl_2}
m <- as.matrix( base::table( dt$logfc_level, dt$md_level) )
m
```


The [spineplot](https://www.rdocumentation.org/packages/graphics/versions/3.6.0/topics/spineplot)
of this contingency table (table(x = logfc_level, y = md_level)) help to visualize the
existent relationship between $|logFC|$ and $MD$.

```{r spineplot_2}
par( mar = c(5, 7, 4, 4), family = "Serif", font = 2, cex = 0.8, las = 1, lwd = 0.2 )
tb <- spineplot(m[, c(3,2,1)], col = c( "blue", "dodgerblue","deepskyblue" ),
                xlab = "", ylab = "")
mtext(side = 1, text = expression(italic(logFC)), 
      line = 2.5, cex = 1, font = 3)
mtext(side = 2, text = expression(paste(italic("MD"))),
      line = 4, cex = 1, font = 3)
abline(a = 0.0, b = 0.72, col = "white", lty = "dashed", lwd = 1)
```

A dashed line was added to emphasize the direction of linear trend. From _"Low"_ to 
_"high"_ there seems to be a linear relationship between pairwise proportion of
$|logFC|$ and $MD$.

Conceptually, the _spineplot_ is a plot of the conditional probability $P(y|x)$
against $P(x)$. Since the joint probability distribution of these variable is
described by a FGM copula, a linear relationship between $P(y|x)$ against $P(x)$
must be expected, which is a property of FGM copula [@Balakrishna2009].

### Testing association between $logFC$ and $MD$ (II)
Here, we use the LLT test implemented in function  [lbl_test](https://www.rdocumentation.org/packages/coin/versions/1.3-0/topics/ContingencyTests)
from the R package _coin_. Two test variants are available: 1) permutation test 
and asymptotic test. 

The asymptotic LLT test:
```{r allt_2}
lbl_test( m )
```


The permutation LLT test:
```{r pllt_2}
lbl_test( m, distribution = "approximate", B = 10000, alternative = "less" )
```


The null hypothesis of no association among the variables in the table is
rejected. In the current case the $p-value < 1e-04$ and, hence, we do not 
have enough evidence to support the null hypothesis.

## Generalized linear model (GLM) (II)

The variable $MD$ is split int three levels, 0:"Low" ($0 < MD \leq 0.5$),
1:"Middle" ($0.5 < MD \leq 1$), and 2:"High" ($1 < MD $)
```{r glm_dt_2}
meth_cut1 <- 0.5
meth_cut2 <- 1.2

xy <- cbind(MD, logFC)
idx <- which(logFC > 0.5)
xy <- data.frame(xy[idx, ])
xy$md <- 0
xy$md[xy$MD > meth_cut1] <- 1
xy$md[xy$MD > meth_cut2] <- 2
```


The glm support the decrease of $-logFC$ with the increment of $MD$. The model
coefficients are higly statistically significant.
```{r glm_2}
gm <- glm(formula = logFC ~ md, data=xy, family = Gamma(link = "identity"))
summary( gm )
```

The ANOVA test confirm the negative trend given by the coefficient of MD 
(highly statistically significant).
```{r anova_gamma_2}
anova(gm, test = "LRT")

```

### GLM group I ("Low" $MD$) versus group II ("Middle" $MD$) (II)

Only genes with $MD < 1$ are analyzed. That is, only two groups are considered:
0:"Middle" ($0.5 < MD \leq 1$), and 1:"High" ($1 < MD $)

```{r glm_3}
dt <- xy[xy$MD < meth_cut2, ]
dt$md <- 0
dt$md[ dt$MD > meth_cut1   ] <- 1

gm <- glm(formula = logFC ~ md, data = dt, family = Gamma(link = "identity"))
summary( gm )

cat("\n =============== The ANOVA test  =============== \n\n")
anova(gm, test = "LRT")

```

### GLM group II ("Middle" $MD$) versus group III ("High" $MD$) (II)
Only genes with $MD < 1$ are analyzed. That is, only two groups are considered:
0:"Low" ($0 < MD \leq 0.5$), 1:"Middle" ($0.5 < MD \leq 1$).


```{r glm_4}
dt <- xy[xy$MD > meth_cut1, ]
dt$md <- 0
dt$md[ dt$MD > meth_cut2   ] <- 1

gm <- glm(formula = logFC ~ md, data = dt, family = Gamma(link = "identity"))
summary( gm )

cat("\n =============== The ANOVA test  =============== \n\n")
anova(gm, test = "LRT")

```

### Boxplot (II)
The boxplot suggest a clear trend of $-logFC$ decreasing with increment of MD
levels. Notice that the _notchs_ from the boxes do not overlap, which implies
that the differences between medians from three groups ("Low", "Middle", and
"High") are statistically significant. So, the GLM results can be visualized.

```{r boxplot_3, fig.width=9, fig.height=6}
par(bty = "n", family = "serif", font = 2, cex = 1, lwd = 0.1)
boxplot(logFC ~ md, data = xy, ylim = c(0.4, 2.5), xaxt ="n",  yaxt = "n", ann = FALSE)
rect(0.5, 0.4, 3.5, 5, col = "grey94", lty = 0, lwd = 1)
grid(0, 5, col = "white", lwd = 2, lty = 1)
boxplot(logFC ~ md, data = xy, ylim = c(0.4, 2.5), notch = TRUE, xaxt ="n",
        yaxt = "n", ann = FALSE, pch = 20, medcol = "darkblue", whisklwd = 1, staplelwd = 1,
        col = c("deepskyblue", "dodgerblue", "dodgerblue3" ), outcol = "red", add  = TRUE)
axis(1, padj = -0.6, at = 1:3, lwd = 0.5, tck = -0.02, line = -0.6, cex.axis = 1.3, 
     labels = c("Low", "Middle", "High"))
axis(2, hadj = 0.6, las = 1, lwd = 0.5, tck = -0.02, line = -1, cex.axis = 1.3)
mtext(side = 1, text = expression(paste(italic("MD"))),
      line = 1.3, cex = 1.4, font = 3)
mtext(side = 2, text = expression(italic(-logFC)), line = 1, cex = 1.4, font = 3)
text(x = 1.5, y = 2.2, "**", pos = 3, cex = 1.2 )

```

# References

<div id="refs"></div>


# Session info {.unnumbered}

Here is the output of `sessionInfo()` on the system on which this document was
compiled running pandoc `r rmarkdown::pandoc_version()`:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```















