---
title: "Methylation analysis with Methyl-IT"
subtitle: <h1>An example with simulated datasets. A guide for users</h>
author: |
    | Robersy Sanchez
    | rus547@psu.edu
    | Mackenzie's lab
    | 
    | Department of Biology and Plant Science. 
    | Pennsylvania State University, University Park, PA 16802
date: "`r format(Sys.time(), '%d %B %Y')`"
fontsize: 11pt
fontfamily: "serif"
bibliography: bibliography.bib
output:
    rmarkdown::html_document: 
        toc: true
        toc_depth: 3
        number_sections: true
        geometry: margin=1in
        keep_md: yes
abstract: |
    Methylation analysis with [Methyl-IT](https://github.com/genomaths/MethylIT)
    is illustrated on simulated datasets of methylated and unmethylated read
    counts with relatively high average of methylation levels: 0.15 and 0.286
    for control and treatment groups, respectively. The main Methyl-IT
    downstream analysis is presented alongside the application of Fisher's exact
    test. The importance of a signal detection step is shown.
vignette: >
    %\VignetteIndexEntry{Methylation analysis with Methyl-IT}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

<style type="text/css">

.list-group-item.active, .list-group-item.active:focus, 
.list-group-item.active:hover {
    z-index: 2;
    color: #fff;
    background-color: #337ab7;
    border-color: #337ab7;
}

.tocify-subheader .tocify-item {
    font-size: 0.80em;
    padding-left: 25px;
    text-indent: 0;
}

.tocify-header .tocify-item {
    font-size: 0.85em;
    padding-left: 20px;
    text-indent: 0;
}

body{ /* Normal  */
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    text-align: justify
}
td {  /* Table  */
    font-size: 8px;
}

h1.title {
    font-size: 38px;
    font-family: "Times New Roman", Times, serif;
    color: DarkRed;
    .subTitle {
        font-size: 24px;
        font-family: "Times New Roman", Times, serif;
        color: DarkRed;
    }
}

h1 { /* Header 1 */
    font-size: 28px;
    font-family: "Times New Roman", Times, serif;
    color: DarkBlue;
}
h2 { /* Header 2 */
    font-size: 22px;
    color: DarkBlue;
    font-family: "Times New Roman", Times, serif;
}
h3 { /* Header 3 */
    font-size: 18px;
    color: DarkBlue;
    font-family: "Times New Roman", Times, serif;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Background
Methyl-IT R package offers a methylome analysis approach based on information
thermodynamics (IT) and signal detection. Methyl-IT approach confront detection
of differentially methylated cytosine as a signal detection problem. This
approach was designed to discriminate methylation regulatory signal from
background noise induced by molecular stochastic fluctuations. Methyl-IT R
package is not limited to the IT approach but also includes Fisher's exact test
(FT), Root-mean-square statistic (RMST) or Hellinger divergence (HDT) tests.
Herein, we will show that a signal detection step is also required for FT, RMST,
and HDT as well. It is worthy to notice that, as for any standard statistical
analysis, any serious robust methylation analysis requires for a descriptive
statistical analysis at the beginning or at different downstream steps. This is
an ABC principle taught in any undergraduate course on statistics. Methylation
analysis is not the exception of the rule. This detail is included in this
example.

Note: This example was made with the MethylIT version at
https://github.com/genomaths/MethylIT. 

# Available datasets and reading
Methylome datasets of whole-genome bisulfite sequencing (WGBS) are available at
Gene Expression Omnibus (GEO DataSets). The data set are downloaded providing
the GEO accession numbers for each data set to the function *getGEOSuppFiles*
(for details type ?getGEOSuppFiles in the R console). Then, datasets can be read
using function *readCounts2GRangesList*. An example on how to load datasets of
read-counts of methylated and unmethylated cytosine into Methyl-IT is given in
the Cancer example at https://git.psu.edu/genomath/MethylIT

# Data generation
For the current example on methylation analysis with Methyl-IT we will use
simulated data. Read-count matrices of methylated and unmethylated cytosine are
generated with Methyl-IT function [simulateCounts](https://is.gd/gg0yTp).

Function *simulateCounts* randomly generates prior methylation levels using Beta
distribution function. The expected mean of methylation levels that we would
like to have can be estimated using the auxiliary function:

```{r Exp-methlevel, include = TRUE}
bmean <- function(alpha, beta) alpha/(alpha + beta)
alpha.ct <- 0.09
alpha.tt <- 0.2
c(control.group = bmean(alpha.ct, 0.5), treatment.group = bmean(alpha.tt, 0.5), 
    mean.diff = bmean(alpha.tt, 0.5) - bmean(alpha.ct, 0.5)) 

```

This simple function uses the $\alpha$ (*shape1*) and $\beta$ (*shape2*) 
parameters from the Beta distribution function to compute the expected value of 
methylation levels. In the current case, we expect to have a difference of 
methylation levels about 0.133 between the control and the treatment.

## Simulation
Function *simulateCounts* from 
[MethylIT.utils](https://github.com/genomaths/MethylIT.utils) 
R package will be used to generate the datasets, which will include three group 
of samples: reference, control, and treatment.

```{r simulation, include = TRUE, message = FALSE}
suppressPackageStartupMessages({
    library(MethylIT)
    library(MethylIT.utils)}
)

# The number of cytosine sites to generate
sites = 50000 
# Set a seed for pseudo-random number generation
set.seed(124)
control.nam <- c("C1", "C2", "C3")
treatment.nam <- c("T1", "T2", "T3")

# Reference group 
ref0 = simulateCounts(num.samples = 4, sites = sites, alpha = alpha.ct, 
                    beta = 0.5, size = 50, theta = 4.5, 
                    sample.ids = c("R1", "R2", "R3", "R4"))
# Control group
ctrl = simulateCounts(num.samples = 3, sites = sites, alpha = alpha.ct, 
                    beta = 0.5, size = 50, theta = 4.5,
                    sample.ids = control.nam)
# Treatment group
treat = simulateCounts(num.samples = 3, sites = sites, alpha = alpha.tt, 
                    beta = 0.5, size = 50, theta = 4.5, 
                    sample.ids = treatment.nam)
```

Notice that reference and control groups of samples are not identical but belong
to the same population.

## Divergences of methylation levels
The estimation of the divergences of methylation levels is required to proceed
with the application of signal detection basic approach. The information
divergence is estimated here using the function *estimateDivergence*. For each
cytosine site, methylation levels are estimated according to the formulas:
$p_i={n_i}^{mC_j}/({n_i}^{mC_j}+{n_i}^{C_j})$, where ${n_i}^{mC_j}$ and
${n_i}^{C_j}$ are the number of methylated and unmethylated cytosines at site 
$i$.

If a Bayesian correction of counts is selected in function *estimateDivergence*,
then methylated read counts are modeled by a beta-binomial distribution in a
Bayesian framework, which accounts for the biological and sampling variations
[@Hebestreit2013; @Robinson2014;
@Dolzhenko2014]. In our case we adopted the Bayesian approach suggested in
reference [@Baldi2001] (Chapter 3).

Two types of information divergences are estimated: *TV*, total variation (*TV*,
absolute value of methylation levels) and Hellinger divergence (*H*). *TV*
is computed according to the formula: $TV_d=|p_{tt}-p_{ct}|$ and *H*:

$$H(\hat p_{ij},\hat p_{ir}) = w_i\Big[(\sqrt{\hat p_{ij}} - \sqrt{\hat
p_{ir}})^2+(\sqrt{1-\hat p_{ij}} - \sqrt{1-\hat p_{ir}})^2\Big]$$ 
where $w_i = 2 \frac{m_{ij} m_{ir}}{m_{ij} + m_{ir}}$, 
$m_{ij} = {n_i}^{mC_j}+{n_i}^{uC_j}+1$, $m_{ir} = {n_i}^{mC_r}+{n_i}^{uC_r}+1$ 
and $j \in {\{c,t}\}$

The equation for Hellinger divergence is given in reference [@Basu2010], but
any other information theoretical divergences could be used as well.

Divergences are estimated for control and treatment groups in respect to a
virtual sample, which is created applying function *poolFromGRlist* on the
reference group.

```{r pooling, include = TRUE}
# Reference sample
ref = poolFromGRlist(ref0, stat = "mean", num.cores = 4L, verbose = FALSE)

# Methylation level divergences
DIVs <- estimateDivergence(ref = ref, indiv = c(ctrl, treat), Bayesian = TRUE, 
                            num.cores = 6L, percentile = 1, verbose = FALSE)

```

The mean of methylation levels differences is:
```{r tv.mean, include = TRUE}
unlist(lapply(DIVs, function(x) mean(mcols(x[, 7])[,1])))
```

# Methylation signal

Likewise for any other signal in nature, the analysis of methylation signal 
requires for the knowledge of its probability distribution. In the current case 
, the signal is represented in terms of the Hellinger divergence of methylation
levels (*H*).

```{r dens, include = TRUE}
# To remove hd == 0 to estimate. The methylation signal only is given for  
divs = lapply(DIVs, function(div) div[ div$hdiv > 0 ], keep.attr = TRUE)
names(divs) <- names(DIVs)

# Data frame with the Hellinger divergences from both groups of samples samples 
l = c(); for (k in 1:length(divs)) l = c(l, length(divs[[k]]))
data_div <- data.frame(H = c(abs(divs$C1$hdiv), abs(divs$C2$hdiv), 
                            abs(divs$C3$hdiv), abs(divs$T1$hdiv), 
                            abs(divs$T2$hdiv), abs(divs$T3$hdiv)),
                    sample = c(rep("C1", l[1]), rep("C2",l[2]), rep("C3",l[3]),
                                rep("T1",l[4]), rep("T2",l[5]), rep("T3",l[6]))
)

```

Empirical critical values for the probability distribution of $H$ and $TV$ can
be obtained using *quantile* function from the R package *stats*.
```{r critical.val, include = TRUE}
critical.val <- do.call(rbind, lapply(divs, function(x) {
    hd.95 = quantile(x$hdiv, 0.95)
    tv.95 = quantile(abs(x$bay.TV), 0.95)
    return(c(tv = tv.95, hd = hd.95))
}))

critical.val

```

The kernel density estimation yields the empirical density shown in the
graphics:

```{r dens.graph, include = TRUE}
suppressMessages(library(ggplot2))

# Some information for graphic
crit.val.ct <- round(max(critical.val[c("C1", "C2", "C3"), 2])) # 67
crit.val.tt <- round(min(critical.val[c("T1", "T2", "T3"), 2])) # 139

# Density plot with ggplot
ggplot(data_div, aes(x = H, colour = sample, fill = sample)) + 
    geom_density(alpha = 0.05, bw = 0.2, position = "identity", na.rm = TRUE,
            size = 0.4) +   
    xlab(expression(bolditalic("Hellinger divergence (H)"))) + 
    ylab(expression(bolditalic("Density"))) +
    ggtitle("Density distribution for control and treatment") +
    geom_vline(xintercept = crit.val.ct, color = "red", 
                linetype = "dashed", size = 0.4) +
    annotate(geom = "text", x = crit.val.ct-2, y = 0.08, size = 5,
            label = 'bolditalic(H[alpha == 0.05]^CT == 67)',
            family = "serif", color = "red", parse = TRUE) +
    geom_vline(xintercept = crit.val.tt, color = "blue", 
                linetype = "dashed", size = 0.4) +
    annotate(geom = "text", x = crit.val.tt -2, y = 0.1, size = 5,
            label = 'bolditalic(H[alpha == 0.05]^TT == 139)',
            family = "serif", color = "blue", parse = TRUE) +
    theme(
        axis.text.x  = element_text( face = "bold", size = 12, color="black",
                                margin = margin(1,0,1,0, unit = "pt" )),
        axis.text.y  = element_text( face = "bold", size = 12, color="black", 
                                margin = margin( 0,0.1,0,0, unit = "mm")),
        axis.title.x = element_text(face = "bold", size = 13,
                                color="black", vjust = 0 ),
        axis.title.y = element_text(face = "bold", size = 13,
                                color="black", vjust = 0 ),
        legend.title = element_blank(),
        legend.margin = margin(c(0.3, 0.3, 0.3, 0.3), unit = 'mm'),
        legend.box.spacing = unit(0.5, "lines"),
        legend.text = element_text(face = "bold", size = 12, family = "serif")
    )

```

The graphic above shows that with high probability the methylation signal 
induced by the treatment has *H* values $H^{TT}_{\alpha=0.05}\geq139$. 
According to the critical value estimated for the differences of methylation 
levels, the methylation signal holds $TV^{TT}_{\alpha=0.05}\geq0.67$. 

Notice that most of the methylation changes are not signal but methylation 
background noise (found to the left of the critical values). This situation is 
typical for all the natural and technologically generated signals 
[@Sanchez2016; @Sanchez2019].

## Signal detection
Signal detection is a critical step to increase sensitivity and
resolution of methylation signal by reducing the signal-to-noise ratio and
objectively controlling the false positive rate and prediction accuracy/risk.

The first estimation in our signal detection step is the identification of the
cytosine sites carrying potential methylation signal $PS$. To identify cytosine 
sites carrying $PS$, we must find the best fitted probability distribution 
model. 

### _Estimation of the best fitted probability distribution model_
Potential DMPs can be estimated using the critical values derived from the best
fitted probability distribution model, which is obtained after the non-linear
fit of the theoretical model on the genome-wide $H$ values for each individual
sample using Methyl-IT function [nonlinearFitDist](https://is.gd/0BLLkT)
[@Sanchez2016]. As before, only cytosine sites with critical values $TV_d>0.926$
are considered DMPs. Notice that, it is always possible to use any other values
of $H$ and $TV_d$ as critical values, but whatever could be the value it will
affect the final accuracy of the classification performance of DMPs into two
groups, DMPs from control and DMPs from treatment (see below). So, it is
important to do an good choices of the critical values.

Function [gofReport](https://is.gd/W7dlP0) search for the best fitted model
between the set of models requested by the user.

```{r nlms, message=FALSE, results='hide'}
d <- c("Weibull2P", "Weibull3P", "Gamma2P", "Gamma3P", "GGamma3P")
gof <- gofReport(HD = DIVs, column = 9L, model = d, num.cores = 4L,
            tasks = 2L, output = "all", verbose = FALSE)
gof$bestModel
```

There is not a one-to-one mapping between $TV_d$ and $H$. However, at each
cytosine site $i$, these information divergences hold the inequality:
$$TV_d(p^{tt}_i,p^{ct}_i)\leq \sqrt{2}H_d(p^{tt}_i,p^{ct}_i)$$

where $H_d(p^{tt}_i,p^{ct}_i)=\sqrt{\frac{H(p^{tt}_i,p^{ct}_i)}w}$ is the
Hellinger distance and $H(p^{tt}_i,p^{ct}_i)$ is the Hellinger divergence
equation given above.

### _Potential methylation signal_
Methylation regulatory signals do not hold the theoretical distribution and,
consequently, for a given level of significance $\alpha$ (Type I error
probability, e.g. $\alpha = 0.05$), cytosine positions $k$ with information
divergence $H_k >= H_{\alpha = 0.05}$ can be selected as sites carrying
potential signals $PS$. The $PS$ is detected with the function 
[getPotentialDIMP](https://is.gd/yGORtg).

```{r getPotentialDIMP}
PS = getPotentialDIMP(LR = DIVs, dist.name = gof$bestModel, nlms = gof$nlms,
                    div.col = 9, alpha = 0.05, tv.col = 8, tv.cut =  0.674)
```

Observe that we have used the $|TV|$ values based on the Bayesian estimation 
of methylation levels (bay.TV, tv.col = 8).

### _Potential DMPs detected with Fisher's exact test (FT)_
Fisher's exact test (FT) is run here as standard alternative approach. A full
discussion on the performance of FT and other statistical approaches in
comparison with Methyl-IT pipeline can be read in [@Sanchez2019]. Notice that
the standard application of FT does not include a signal detection step (see
more at [@Sanchez2019]).

In Methyl-IT, FT is implemented in function *FisherTest*. In the current case, a
pairwise group application of FT to each cytosine site is performed. The
differences between the group means of read counts of methylated and
unmethylated cytosines at each site are used for testing ($pooling.stat
="mean"$). Notice that only cytosine sites with critical values $TV_d>0.67$ are
tested (*tv.cut = 0.67*).

```{r FT, include = TRUE, message=FALSE, results='hide'}
ft = FisherTest(LR = divs, tv.cut = 0.67,
                    pAdjustMethod = "BH",  pooling.stat = "mean", 
                    pvalCutOff = 0.05, num.cores = 4L,
                    verbose = FALSE, saveAll = FALSE) 

# TV column is the standard TV value (without Bayesian correction of methylation
# levels)
ft.tv <- getPotentialDIMP(LR = ft, div.col = 9L, dist.name = "None",
                        tv.cut = 0.67, tv.col = 7, alpha = 0.05)

```

The above setting would impose additional constrains on the output 
of DMPs resulting from Fisher's exact test depending on how strong is the
methylation background noise in the control  population. Basically, methylation 
variations that can spontaneously occur in the control population with 
relatively high frequencies are disregarded. The decisions are based on the 
empirical cumulative distribution function (ECDF).


Potential DMPs detected with FT can be constrained with the critical values
derived for the best fitted probability distribution models

```{r FT.hd, include = TRUE}
# Potential DMPs from Fisher's exact test 
ft.hd <- getPotentialDIMP(LR = ft, div.col = 9L, nlms = gof$nlms, 
                        tv.cut = 0.674, tv.col = 8, alpha = 0.05, 
                        dist.name = gof$bestModel)
ft.hd$T1
```

The above result shows column 'adj.pval' with the FT's p.values adjusted for 
multiple comparisons and column 'wprob' with the predicted probabilities: 
$1 - P(H_k >= H_{\alpha = 0.05})$ from the best fitted model.

Summary table:
```{r FT.ecdf, include = TRUE}
data.frame(ft = unlist(lapply(ft.tv, length)),
            ft.hd = unlist(lapply(ft.hd, length)),
BestModel = unlist(lapply(PS, length)))

```

Notice that there are significant differences between the numbers of PSs
detected by FT and other approaches. There at least about 7800 - 2226 = 5574 or
more PSs in the treatment samples that are not induced by the treatment, which
can naturally occurs in both groups, control and treatment. In other words, the
best probabilistic distribution models suggest that those methylation events are
resultant of the normal (non-stressful) biological processes. Any information
divergence of methylation level below the critical value $TV_{\alpha=0.05}$ or
$H_{\alpha=0.05}$ is a methylation event that can occur in normal conditions.
*It does not matter how much significant a test like FT could be, since it is
not about the signification of the test, but about how big is the probability to
observe that methylation event in the control population as well*.

### Critical values based on fitted models
Critical values $H_{\alpha=0.05}$ can be estimated from each fitted model using
[predict](https://is.gd/CCqNYS) function:

```{r pred}
# ==== 95% quantiles 
q95 <- data.frame(q95 = predict(nlm = gof$nlms, pred = "quant", q = 0.95,
                                dist.name = gof$bestModel))
q95
```

# Cutpoint for the spontaneous variability in the control group 
Normally, there is a spontaneous variability in the control group. This is a
consequence of the random fluctuations, or noise, inherent to the methylation
process. The stochasticity of the the methylation process is derives from the
stochasticity inherent in biochemical processes. There are fundamental physical
reasons to acknowledge that biochemical processes are subject to noise and
fluctuations [@Samoilov2006; @Eldar2010]. So, regardless constant environment,
statistically significant methylation changes can be found in control population
with probability greater than zero and proportional to a Boltzmann factor
[@Sanchez2016].

Natural signals and those generated by human technology are not free of noise
and, as mentioned above, the methylation signal is no exception. Only signal
detection based approaches are designed to filter out the signal from the noise,
in natural and in human generated signals.

The need for the application of (what is now known as) signal detection in
cancer research was pointed out by Youden in the midst of the last century
[@Youden1950]. Here, the application of signal detection approach was performed
according with the standard practice in current implementations of clinical
diagnostic test [@Carter2016; @lopez2014; @Hippenstiel2001]. That is, optimal
cutoff values of the methylation signal were estimated on the receiver operating
characteristic curves (ROCs) and applied to identify DMPs. The decision of
whether a DMP detected by Fisher’s exact test (or any other statistical test
implemented in Methyl-IT) is taken based on the optimal cutoff value.

In the current example, the column carrying *TV* (*div.col* = 7L) will be used
to estimate the cutpoint. The column values will be expressed in terms of 
$TV_d=|p_{tt}-p_{ct}|$ (*absolute* = TRUE):


### Cutpoint for PS estimated based on the best probability distribution model
Next, the search for an optimal cutpoint is accomplished with a model classifier
and using the total variation distance (with Bayesian correction, bay.TV)
```{r cutpoints}
cutpoint = estimateCutPoint(LR = PS, 
                            control.names = c("C1","C2", "C3"), 
                            treatment.names = c("T1", "T2", "T3"),
                            simple = FALSE,
                            classifier1 = "pca.logistic", 
                            classifier2 = "pca.qda", 
                            column = c(hdiv = TRUE, bay.TV = TRUE, 
                                        wprob = TRUE, pos = TRUE),
                            n.pc = 4 , center = TRUE, scale = TRUE,
                            div.col = 9, clas.perf = TRUE)
cutpoint$cutpoint
cutpoint$testSetPerformance
cutpoint$testSetModel.FDR

```


### Cutpoint for PS estimated based on FT
The best classification performance based PS estimated based on FT are obtained
with the application of [Youden index](https://is.gd/t8UKvK) [@Youden1950].

```{r cutpoints2}
cutpoint2 = estimateCutPoint(LR = ft.tv, 
                            control.names = c("C1","C2", "C3"), 
                            treatment.names = c("T1", "T2", "T3"),
                            simple = TRUE,
                            classifier1 = "pca.qda", 
                            column = c(hdiv = TRUE, bay.TV = TRUE, 
                                        wprob = TRUE, pos = TRUE),
                            n.pc = 4 , center = TRUE, scale = TRUE,
                            div.col = 9, clas.perf = TRUE,
                            post.cut = 0.9)
cutpoint2$cutpoint
cutpoint2$testSetPerformance
cutpoint2$testSetModel.FDR

```

Addition of information on the probability distribution of the methylation
signal to potential signal identified with FT will emulate the results obtained
with the Methyl-IT pipeline. 

```{r cutpoints3}
cutpoint3 = estimateCutPoint(LR = ft.hd, 
                            control.names = c("C1","C2", "C3"), 
                            treatment.names = c("T1", "T2", "T3"),
                            simple = FALSE,
                            classifier1 = "pca.logistic", 
                            classifier2 = "pca.qda", 
                            column = c(hdiv = TRUE, bay.TV = TRUE, 
                                        wprob = TRUE, pos = TRUE),
                            n.pc = 4 , center = TRUE, scale = TRUE,
                            div.col = 9, clas.perf = TRUE)
cutpoint3$cutpoint
cutpoint3$testSetPerformance
cutpoint3$testSetModel.FDR

```

Therefore, invoking the parsimony principle, we assume that signal detection and
machine-learning classifiers are sufficient [@ElNaqa2018].

## DMPs
Cytosine sites carrying a methylation signal are designated differentially
informative methylated positions (DMPs). With high probability true DMPs can be
selected with Methyl-IT function [selectDIMP](https://is.gd/7cVEEd). 
```{r DMPs}
DMPs = selectDIMP(PS, div.col = 9, cutpoint = cutpoint$cutpoint)

```

### Evaluation of DMP classification
As shown above, DMPs are found in the control population as well. Hence, it is
important to know whether a DMP is the resulting effect of the treatment or just
spontaneously occurred in the control sample as well. In particular, the
confrontation of this issue is extremely important when methylation analysis is
intended to be used as part of a diagnostic clinical test and a decision making
in biotechnology industry.

Methyl-IT function
[evaluateDIMPclass](https://is.gd/nzy9j3)
is used here to evaluate the classification of DMPs into one of the two
classes, control and treatment. Several classifiers are available to be used
with this function (see the help/manual for this function or type
?evaluateDIMPclass in command line).

A first evaluation of DMP classification performance was already done after
setting "clas.perf = TRUE" in function
[estimateCutPoint](https://is.gd/RR9iWY).
However, this performance derives just from randomly split the set of sample
into two subsets: training (60%) and testing (40%). A good classification
performance report would be just the result from a lucky split of the set of
samples! Hence, what would happen if the random split and further analysis are
repeated 300 times?

```{r g2p.class, include = TRUE}
class.perf = evaluateDIMPclass(LR = DMPs, control.names = control.nam,
                            treatment.names = treatment.nam,
                            column = c(hdiv = TRUE, TV = TRUE, 
                                        wprob = TRUE, pos = TRUE),
                            classifier = "pca.qda", n.pc = 4, pval.col = 10L, 
                            center = TRUE, scale = TRUE, num.boot = 300,
                            output = "mc.val", prop = 0.6, num.cores = 4L,
                            tasks = 2L
)
class.perf
```

# Conclusions Summary
Herein, an illustrative example of methylation analysis with Methyl-IT have been
presented. Whatever could be the statistical test/approach used to identify
DMPs, the analysis with simulated datasets, where the average of methylation
levels in the control samples is relatively high, indicates the need for the
application of signal detection based approaches.

For the current simulated dataset, the best classification performance was
obtained for the approach of DMP detection based on the best fitted probability
distribution model for the Hellinger divergence of methylation levels. DMPs
from treatment are distinguish from control DMPs with very high accuracy.

## Concluding remarks
The simplest suggested steps to follow for a methylation analysis with Methyl-IT
are:

\item 1. To estimate a reference virtual sample from a reference group by using
function [poolFromGRlist](https://is.gd/qR79kW). 
Notice that several statistics are available to
estimate the virtual samples, i.e., *mean*, *median*, *sum*. For experiments
limited by the number of sample, at least, try the estimation of the virtual
sample from the control group. Alternatively, the whole reference group can be 
used in pairwise comparisons with control and treatment groups (computationally
expensive).

\item 2. To estimate information divergence using function
[estimateDivergence](https://is.gd/0ViBO9)

\item 3. To perform the estimation of the best fitted probability distribution
model using function [gofReport](https://is.gd/W7dlP0).

\item 4. To get the potential DMPs using function 
[getPotentialDIMP](https://is.gd/yGORtg).

\item 5. To estimate the optimal cutpoint using function
[estimateCutPoint](https://is.gd/RR9iWY).

\item 6. To retrieve DMPs with function [selectDIMP](https://is.gd/7cVEEd).

\item 7. To evaluate the classification performance using function 
[evaluateDIMPclass](https://is.gd/nzy9j3).

# References 













