---
title: "Get started-with Methyl-IT"
author: 
    - name: Robersy Sanchez
      affiliation: Department of Biology.
                   Pennsylvania State University, University Park, PA 16802
      email: rus547@psu.edu
date: "`r format(Sys.time(), '%d %B %Y')`"
fontfamily: "serif"
bibliography: bibliography.bib
output: 
    BiocStyle::html_document:
        toc: true
        toc_float: true
        toc_depth: 2
package: BiocStyle
abstract: |
    A fast introduction into methylation analysis with Methyl-IT is provided
    here. The methylome analysis implemented in Methyl-IT based on a
    signal-detection and machine-learning approach. Vignettes with 
    further examples are available at <https://genomaths.github.io/methylit/>.
vignette: >
    %\VignetteIndexEntry{Get started-with Methyl-IT}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::knitr}
---

<style type="text/css">

.list-group-item.active, .list-group-item.active:focus, 
.list-group-item.active:hover {
    z-index: 2;
    color: #fff;
    background-color: #337ab7;
    border-color: #337ab7;
}

.tocify-subheader .tocify-item {
    padding-left: 25px;
    text-indent: 0;
}

.tocify-header .tocify-item {
    padding-left: 20px;
    text-indent: 0;
}

body{ /* Normal  */
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    text-align: justify
}

h1.title {
    font-family: "Times New Roman", Times, serif;
    color: DarkRed;
    .subTitle {
        font-family: "Times New Roman", Times, serif;
        color: DarkRed;
    }
}

h1 { /* Header 1 */
    font-family: "Times New Roman", Times, serif;
    color: DarkBlue;
}
h2 { /* Header 2 */
    color: DarkBlue;
    font-family: "Times New Roman", Times, serif;
}
h3 { /* Header 3 */
    color: DarkBlue;
    font-family: "Times New Roman", Times, serif;
}
code.r{ /* Code block */
    font-size: 14px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
.sidebar {
font-size: 9px;
text-indent: 0;
}

</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```


```{r style, echo = FALSE, results = 'asis'}
  BiocStyle::markdown()
```


```{r setup, message=FALSE}
library(MethylIT)
```

# Background
MethylIT is an R package for methylome analysis based on a signal-detection
machine learning approach (SD-ML). This approach is postulated to provide
greater sensitivity for resolving true signal from the methylation background
within the methylome [@Sanchez2016; @Sanchez2019]. Because the biological signal
created within the dynamic methylome environment characteristic of plants is not
free from background noise, the approach, designated Methyl-IT, includes the
application of signal detection theory [@Greiner2000; @Carter2016; @Harpaz2013;
@Kruspe2017]. A basic requirement for the application of signal detection is the
knowledge of the probability distribution of the background noise, which is used
as null hypothesis in the detection of the methylation signal. This probability
distribution is, in general, a member of generalized gamma distribution family,
and it can be deduced on a statistical mechanical/thermodynamics basis from DNA
methylation induced by thermal fluctuations [@Sanchez2016].

# Data sets

Methyl-IT uses the read counts of methylation called derived from whole-genome
bisulfite sequencing (WGBS). Methylome datasets of WGBS are available at Gene
Expression Omnibus (GEO DataSets). For the current example, we just load RData
files containing the read-counts from several samples. 

```{r reads, message=FALSE}
## ===== Download methylation data from PSU's GitLab ========
url <- paste0("https://git.psu.edu/genomath/MethylIT_examples",
            "/raw/master/MethylIT_testing/",
            "memoryLine_100k_samples_cg.RData")
temp <- tempfile(fileext = ".RData")
download.file(url = url, destfile = temp)
load(temp)
file.remove(temp); rm(temp, url)

names(meth_samples)

```

## Gene annotation
Download _Arabidopsis thaliana_ TAIR10 gene annotation data
```{r genes, message=FALSE}
url <- paste0("https://git.psu.edu/genomath/MethylIT_examples",
            "/raw/master/MethylIT_testing/",
            "Arabidopsis_thaliana.TAIR10.38_only_genes.RData")
temp <- tempfile(fileext = ".RData")
download.file(url = url, destfile = temp)
load(temp)
file.remove(temp); rm(temp, url)

```

## The reference sample
To perform the comparison between the uncertainty of methylation levels from
each group of individuals, control $(c)$ and treatment $(t)$, we should estimate
the uncertainty variation with respect to the same reference individual on the
mentioned metric space. The reason to measure the uncertainty variation with
respect to the same reference resides in that even sibling individuals follow an
independent ontogenetic development and, consequently, their developments follow
independent stochastic processes. The reference individual is estimated with 
function [poolFromGRlist](https://is.gd/qR79kW):
```{r ref}
ref <- poolFromGRlist(list(meth_samples$W_3_1, meth_samples$W_3_2,
                        meth_samples$W_3_3), stat = "mean", num.cores = 4L
)
```

# Estimation of information divergences
To perform the comparison between the uncertainty of methylation levels
from each group of individuals, control $(c)$ and treatment $(t)$, the
divergence between the methylation levels of each individual is estimated with
respect to the reference sample.

```{r div}
ml_samples <- meth_samples[c("W_3_3", "W_3_4","W_3_5",
                            "M_3_1","M_3_2","M_3_3","M_3_4")]
ml_hd <- estimateDivergence(ref = ref,
                            indiv = ml_samples,
                            Bayesian = TRUE,
                            min.coverage = 4,
                            min.meth = 1,
                            high.coverage = 500,
                            percentile = 0.999,
                            num.cores = 4L,
                            tasks = 2L, verbose = FALSE)
ml_hd$M_3_1
```

Function [estimateDivergence](https://is.gd/0ViBO9) returns a list of GRanges
objects with the four columns of counts, the information divergence, and
additional columns:

1. The original matrix of methylated ($c_i$) and unmethylated ($t_i$) read
   counts from control ($i=1$) and treatment ($i=2$) samples.
2. "p1" and "p2": methylation levels for control and treatment, 
    respectively.
3. "bay.TV": total variation $TV = p_2 - p_1$.
4. "TV": total variation based on simple counts: 
    $TV=c_1/(c_1+t_1)-c_2/(c_2+t_2)$.
5. "hdiv": Hellinger divergence.

Observe that the absolute values $|TV|$ is an information divergence as well,
which is known as total variation distance (widely used in information theory).

## _Descriptive statistical analysis_
A descriptive analysis on the distribution of Hellinger divergences of
methylation levels is recommended. Like in any other statistical analysis,
descriptive statistical analysis help us to detect potential issues in the raw
data. It is the user responsibility to perform quality check of his/her dataset
before start to apply Methyl-IT. Nevertheless, the quality checking of the raw
data is not perfect. So, it is healthy to sees for potential issues.

```{r crit}
## Critical values
critical.val <- do.call(rbind, lapply(ml_hd, function(x) {
    hd.95 = quantile(x$hdiv, 0.95)
    tv.95 = quantile(abs(x$bay.TV), 0.95)
    btv.95 = quantile(abs(x$TV), 0.95)
    return(c(tv = tv.95, hd = hd.95, TTV = btv.95,
            num.sites.hd95 = sum(x$hdiv > hd.95),
            num.sites.tv95 = sum(x$bay.TV > tv.95),
            num.sites.btv.95 = sum(x$TV > btv.95)
    ))})
)
critical.val

```

## Quality checking of the coverage
After the raw data filtering introduced with
[estimateDivergence](https://is.gd/0ViBO9) function, some coverage issue could
be detected.

```{r cov}
covr <- lapply(ml_hd, function(x) {
                cov <- x$c2 + x$t2
                return(cov)
})

do.call(rbind, lapply(covr, function(x) {
    q60 <- quantile(x, 0.6)
    q9999 <- quantile(x, 0.9999)
    idx1 <- which(x >= q60)
    idx2 <- which(x <= 500)
    q95 <- quantile(x, 0.95)
    idx <- intersect(idx1, idx2)
    return(c(
        round(summary(x)),
        q60, quantile(x, c(0.95, 0.99, 0.999, 0.9999)),
        num.siteGreater_8 = sum(x >= 8),
        q60_to_500 = sum((x >= q60) & (x <= 500)),
        num.siteGreater_500 = sum(x > 500)
))})
)
```

# Estimation of the best fitted probability distribution model
A basic requirement for the application of signal detection is the knowledge of
the probability distribution of the background noise. Probability distribution,
as a Weibull distribution model, can be deduced on a statistical
mechanical/thermodynamics basis for DNA methylation induced by thermal
fluctuations [@Sanchez2016].

```{r gof, message=FALSE, results='hide'}
gof <- gofReport(HD = ml_hd,
                model = c("Weibull2P", "Weibull3P",
                        "Gamma2P", "Gamma3P"),
                column = 9,
                output = "all",
                confl_model = TRUE,
                num.cores = 4L,
                task = 2L,
                verbose = FALSE
)
gof$stats
gof$bestModel
```

# Signal detection
The information thermodynamics-based approach is postulated to provide greater
sensitivity for resolving true signal from the thermodynamic background within
the methylome [@Sanchez2016]. Since the biological signal created within the
dynamic methylome environment characteristic of living organisms is not free
from background noise, the approach, designated Methyl-IT, includes the
application of signal detection theory [@Greiner2000; @Carter2016; @Harpaz2013;
@Kruspe2017]. Signal detection is a critical step to increase sensitivity and
resolution of methylation signal by reducing the signal-to-noise ratio and
objectively controlling the false positive rate and prediction accuracy/risk.

## _Potential methylation signal_
The first estimation in our signal detection step is the identification of the
cytosine sites carrying potential methylation signal $PS$. The methylation
regulatory signal does not hold the theoretical distribution and, consequently,
for a given level of significance $\alpha$ (Type I error probability, e.g.
$\alpha = 0.05$), cytosine positions $k$ with information divergence $H_k >=
H_{\alpha = 0.05}$ can be selected as sites carrying potential signals $PS$. The
value of $\alpha$ can be specified. For example, potential signals with $H_k >
H_{\alpha = 0.01}$ can be selected. For each sample, cytosine sites are selected
based on the corresponding fitted theoretical distribution model estimated in
the previous step.

```{r ps}
ps <- getPotentialDIMP(LR = ml_hd,
                        nlms = gof$nlms,
                        div.col = 9L,
                        tv.col = 0.25,
                        dist.name = gof$bestModel)
ps$M_3_1
```

## Cutpoint estimation. DMPs
Laws of statistical physics can account for background methylation, a response
to thermal fluctuations that presumably functions in DNA stability
[@Sanchez2016; @Sanchez2019]. True signal is detected based on the optimal
cutpoint [@lopez2014].

The need for the application of (what is now known as) signal detection in
cancer research was pointed out by Youden in the midst of the last century
[@Youden1950]. In the next example, the simple cutpoint estimation available in
Methyl-IT is based on the application of [Youden
index](https://en.wikipedia.org/wiki/Youden%27s_J_statistic) [@Youden1950].
Although cutpoints are estimated for a single variable, the classification
performance can be evaluated for several variables and applying different model
classifiers. A optimal cutpoint distinguishes disease stages from healthy
individual. The performance of this classification is given in the output of
function [estimateCutPoint](https://is.gd/RR9iWY).

A model classifier can be requested for further predictions and its
classification performance is also provided. Below, the selected model
classifier is a quadratic discriminant analysis (QDA) (*classifier1* = "qda",
*clas.perf* = TRUE). Four predictor variables are available: the Hellinger
divergence of methylation levels (*hdiv*), total variation distance (*TV*,
absolute difference of methylation levels), relative position of cytosine site
in the chromosome (*pos*), and the logarithm base two of the probability to
observe a Hellinger divergence value $H$ greater than the critical value
$H_{\alpha = 0.05}$ (values given as probabilities in object PS, *wprob*). 

Notice that the cutpoint can be estimated for any of the two currently available
information divergences: *Hellinger divergence* (*div.col* = 9) or the *total*
*variation distance* (with Bayesian correction, *div.col* = 8).

```{r cutp}
### Cutpoint estimated using a model classifier
cutpoint <- estimateCutPoint(LR = ps,
                            simple = FALSE,
                            div.col = 9,
                            control.names = c("W_3_3", "W_3_4", "W_3_5"),
                            treatment.names = c("M_3_1", "M_3_2", "M_3_3",
                                                "M_3_4", "M_3_5"),
                            column = c(hdiv = TRUE, bay.TV = TRUE,
                                        wprob = TRUE, pos = TRUE),
                            classifier1 = "pca.qda", n.pc = 4,
                            classifier2 = "pca.logistic",
                            center = TRUE,
                            scale = TRUE,
                            clas.perf = TRUE,
                            verbose = FALSE
)
cutpoint$cutpoint
cutpoint$testSetPerformance
cutpoint$testSetModel.FDR
```


Differentially methylated positions (DMPs) can be now identified using 
[selectDIMP](https://is.gd/7cVEEd) function

```{r dmps}
dmps <- selectDIMP(ps, div.col = 9, cutpoint = cutpoint$cutpoint)
data.frame(dmps =unlist(lapply(dmps, length)))

```

# Differentially methylated genes (DMGs)
Differentially methylated
genes (DMGs) are estimated from group comparisons for the number of DMPs on
gene-body regions between control and treatment. 

Function [getDIMPatGenes](https://is.gd/WQzX5I) is used to count the number of
DMPs at gene-body. Nevertheless, it can be used for any arbitrary specified
genomic region as well. The operation of this function is based on the
'*findOverlaps*' function from the '*GenomicRanges*' Bioconductor R package. The
'*findOverlaps*' function has several critical parameters like, for example,
'*maxgap*', '*minoverlap*', and '*ignore.strand*'. In our function
[getDIMPatGenes](https://is.gd/WQzX5I), except for setting ignore.strand = TRUE
and *type = "within"*, we preserve the rest of default 'findOverlaps'
parameters. In this case, these are important parameter settings because the
local mechanical effect of methylation changes on a DNA region where a gene is
located is independent of the strand where the gene is encoded. That is,
methylation changes located in any of the two DNA strands inside the gene-body
region will affect the flexibility of the DNA molecule [@Choy2010;
@Severin2011].

```{r dmps_at_genes}
nams <- names(dmps)
dmps_at_genes <- getDIMPatGenes(GR = dmps, GENES = genes, ignore.strand = TRUE)
dmps_at_genes <- uniqueGRanges(dmps_at_genes, columns = 2L,
                                ignore.strand = TRUE, type = "equal")
colnames(mcols(dmps_at_genes)) <- nams
dmps_at_genes

```

## _Setting the experiment design_

The experimental design is set with
[glmDataSet](https://genomaths.github.io/methylit/reference/glmDataSet.html) 
function
```{r exper}
colData <- data.frame(condition = factor(c("WT", "WT", "WT",
                                            "ML", "ML", "ML", "ML"),
                                        levels = c("WT", "ML")),
                    nams,
                    row.names = 2)
## A RangedGlmDataSet is created
ds <- glmDataSet(GR = dmps_at_genes, colData = colData)
```

## Testing for DMGs
DMGs are detected using function
[countTest2](https://genomaths.github.io/methylit/reference/countTest2.html).
```{r }
dmgs <- countTest2(DS = ds, num.cores = 4L,
                    tasks = 2L,
                    minCountPerIndv = 7,
                    maxGrpCV = c(1.2, 1.2),
                    Minlog2FC = 0.5,
                    CountPerBp = 0.001,
                    test = "LRT",
                    verbose = TRUE)
dmgs
```

# References

# Session info {.unnumbered}

Here is the output of `sessionInfo()` on the system on which this document was
compiled running pandoc `r rmarkdown::pandoc_version()`:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

