---
title: "Some tips for Methyl-IT R scripts"
author: |
     | Robersy Sanchez
     | rus547@psu.edu
     | Mackenzie's lab
     |
     | Department of Biology and Plant Science. 
     | Pennsylvania State University, University Park, PA 16802
     | Maintainer: Robersy Sanchez
date: "`r format(Sys.time(), '%d %B %Y')`"
fontsize: 11pt
fontfamily: "serif"
output:
    rmarkdown::word_document: 
        toc: true
        toc_depth: 3
        keep_md: yes

vignette: >
    %\VignetteIndexEntry{Principal Components and Linear Discriminant Analyses with Methyl-IT}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}  

---


<style type="text/css">

.list-group-item.active, .list-group-item.active:focus, 
.list-group-item.active:hover {
    z-index: 2;
    color: #fff;
    background-color: #337ab7;
    border-color: #337ab7;
}

.tocify-subheader .tocify-item {
    /* font-size: 0.80em; */ 
    padding-left: 25px;
    text-indent: 0;
}

.tocify-header .tocify-item {
    /* font-size: 0.85em; */ 
    padding-left: 20px;
    text-indent: 0;
}

body{ /* Normal  */
    /* font-size: 18px; */
    font-family: "Times New Roman", Times, serif;
    text-align: justify
}
td {  /* Table  */
    /* font-size: 8px; */
}

h1.title {
    /* font-size: 38px; */
    font-family: "Times New Roman", Times, serif;
    color: DarkRed;
    .subTitle {
        /* font-size: 24px; */
        font-family: "Times New Roman", Times, serif;
        color: DarkRed;
    }
}

h1 { /* Header 1 */
    /* font-size: 28px; */
    font-family: "Times New Roman", Times, serif;
    color: DarkBlue;
}
h2 { /* Header 2 */
    /* font-size: 22px; */
    color: DarkBlue;
    font-family: "Times New Roman", Times, serif;
}
h3 { /* Header 3 */
    /* font-size: 18px; */
    color: DarkBlue;
    font-family: "Times New Roman", Times, serif;
}
code.r{ /* Code block */
    font-size: 14px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reading data sets

Let's suppose that we have the following read-count data set obtained from 
Bismark 'CX_report' files:

```{r files, eval=FALSE}
folder <- "~/data/"
files <- list.files(path = folder)
# [1] "~/data/Col_8-1a.fq_bismark_bt2.deduplicated.bismark.CX_report"
# [2] "~/data/Col_8-1b.fq_bismark_bt2.deduplicated.bismark.CX_report"
# [3] "~/data/Col_8-1c.fq_bismark_bt2.deduplicated.bismark.CX_report"

```

This data set can be read with the following code:

```{r reading, eval=FALSE}
library(MethylIT)
col0 <- readCounts2GRangesList(filenames = paste0(folder, files),
                               sample.id = c("col1", "col2", "col3"),
                               pattern = "^[1-5]",
                               columns = c(seqnames = 1, start = 2,
                                           strand = 3, mC = 4, uC = 5,
                                           context = 6))
```

Since the methylome data are from Arabidopsis thaliana (a plant), we have added
the column 6 from each Bismark 'CX_report' file, which for each cytosine site
carries the information of its corresponding methylation context: CG, CHG, or 
CHH. Users familiar with Linux OS can easily check which is the number of each 
required column by using Linux command line ‘head’ function.
 
```{bash linux, eval=FALSE}
head BS-10.CX_report.txt
1       1       +       0       0       CHH     CCC
1       2       +       0       0       CHH     CCT
1       3       +       0       0       CHH     CTA
1       8       +       0       0       CHH     CCC
1       9       +       0       0       CHH     CCT
1       10      +       0       0       CHH     CTA
1       15      +       0       0       CHH     CCC
1       16      +       0       0       CHH     CCT
1       17      +       0       0       CHH     CTA
1       22      +       0       0       CHH     CCC
```

Plants methylome contains chloroplast and mitochondrial chromosomes, which are
not included in methylome analysis. In the current case, we can exclude these
chromosomes by setting:

```{r p}
pattern = "^[1-5]"
```

That is, only rows starting with any of the integer numbers from 1 to 5 are
read. Notice that [readCounts2GRangesList](https://is.gd/5pyk1N) function is
able to deal with different file formats. The user just need to provide the
number of the specified columns, which for different aligner programs can be
different.

It is very important to notice that Methyl-IT pipeline assumes that all the data
provided have acceptable quality. It is up to the user to guaranty the quality
of its methylome data set. For example, to include chloroplast chromosome, will
artificially decrease the levels of noise in the treatment and control group of
samples, which definitely will lead to a wrong estimation of the cutpoints in
the signal detection step.

### Analysis per methylation context

Usually, in plant epigenomic field, the methylation analysis is performed by
methylation context. After apply the script above, the three methylation 
contexts were included in each sample. The 
[readCounts2GRangesList](https://is.gd/5pyk1N) function can be used to read only
one specified methylation context, but it also increase the number of files per
sample.

The next code will select the methylation context that want, say "CHG", from
the above samples:

```{r chg, eval=FALSE}
col0_chg <- lapply(col0, function(s) s[s$context == "CHG"], keep.attr = TRUE)
```

Next, to compute the reference sample, keep in mind that we have two columns
carrying integer numbers and third column carrying a character vector, which 
will be not included in the computation. To prevent error we use 'columns = 1:2'

```{r ref, eval=FALSE}
ref_chg <- poolFromGRlist(LR = col0_chg, stat = "sum", columns = 1:2,
                          num.cores = 10L)

```

## DMPs

Let's suppose that we have set of DMPs estimated for each methylation context:
dmps_cg, dmp_chg, and dmps_chh. For DMG or DMR analyses we usually include all
the methylation contexts.

```{r dmps, eval=FALSE}
nams <- names(dmps_cg)
dmps <- dmps_cg
for (k in 1:length(nams)) {
  dmps[[k]] <- c(dmps_cg[[k]], dmps_chg[[k]], dmps_chh[[k]])
}
names(dmps) <- nams
```
## Annotation file 

Gene annotation is needed for DMG estimation. Some organisms (not well annotated
yet) reports several "contig" sequences which are not included between the known
chromosomes. In the most frequent case, as in the case of Arabidopsis, we do
not include chloroplast and mitochondrial chromosomes. For Arabidopsis,
following code does job when the "gtf" is used:

```{r annot, eval=FALSE}
library(rtracklayer)
AG = import("~/data/TAIR10_gff3/Arabidopsis_thaliana.TAIR10.46.gtf.gz")
gene = AG[ AG$type == "gene", c( "gene_id", "gene_biotype", "gene_name" ) ]
gene = gene[ gene$gene_biotype == "protein_coding", c("gene_id", "gene_name") ]
seqlevels(gene, pruning.mode = "coarse") <- c("1", "2", "3", "4", "5")
gene = sortBySeqnameAndStart(gene)
```

In fact, the critical step in the above code is:

```{r crit, eval=FALSE}
seqlevels(gene, pruning.mode = "coarse") <- c("1", "2", "3", "4", "5")
```


## General suggestions

1. Theoretically, the nonlinear fit to estimate the probability distribution of 
   the information divergence and the cutpoint estimation must be done per
   chromosome. However, for organisms with small methylome, as in the case of
   Arbidopsis, we have not detected significant differences, so far, when the 
   methylome reprogramming is genome-wide. Nevertheless, if you suspect that the
   methylation reprogramming is affecting specific chromosomes, then we strongly
   suggest to perform the analysis per chromosome. In the case of human 
   methylome, the analysis per chromosome is the right way to proceed.
   
2. The cutpoint estimation step is critical. The evaluation of the 
   classification performance is a very important step. Typically, a 
   classification performance with accuracy below 95% would indicate that 
   either something is wrong in some previous step (not necessarily in 
   Methyl-IT steps) or that really there are not too much difference between 
   your control and treatment groups, as would be the case, for example, in the 
   analysis of different seed development stages. Please read and be familiar 
   with the vignette:  
   [Optimal cutpoint for the methylation signal](https://is.gd/WoYFUe).
   
   




